---
title: "Cascading effects of regime shifts in social-ecological systems"
author: "Juan C. Rocha, Garry Peterson, Örjan Bodin, Simon Levin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document:
    toc: no
  pdf_document:
    dev: pdf
    keep_tex: yes
    latex_engine: pdflatex
    toc: no
citecolor: blue
corresponding_author:
- code: 2
  text: 'To whom correspondence should be addressed. E-mail: juan.rocha@su.se'
csl: nature.csl
address:
- address: Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691
    Stockholm
  code: a
- address: Beijer Institute, Swedish Royal Academy of Sciences, Lilla Frescativägen
    4A, 104 05 Stockholm
  code: b
documentclass: article
fontsize: 9pt
header-includes:
- \usepackage{dcolumn}
- \usepackage{lineno}
- \linenumbers
keywords: regime shifts, cascading effects, tele-connections, coupled critical transitions,
  networks
lineno: yes
linkcolor: blue
bibliography: cascading.bib
abstract: |
  Regime shifts in nature and society are likely to occur more often and severely as humans increase their pressure on the world's ecosystems. Yet it is largely unknown how these transitions will interact, whether the occurrence of one will increase the likelihood of another, or whether they might simply be correlated at distant places. Here we explore three types of potential cascading effects of regime shifts$:$ *i)* two regime shifts can correlate if they are affected by the same driving forces [@Rocha:2015du]; *ii)* a domino effect is triggered when the occurrence of one regime shift affects the drivers of other regime shifts [@Rocha:2010vv; @Hughes:2013cv; @Scheffer:2012cta]; or *iii)* hidden feedbacks rise when two regime shifts combined generate new (not previously identified) feedbacks [@Liu:2015go] by reinforcing or damping each other's drivers [@Rocha:2010vv]. Using 30 regime shifts described as networks, we show that 80% of the pair-wise combinations of regime shifts present at least one cascading effect, and 45% present plausible structural interdependence. Driver sharing is more common in aquatic systems, while hidden feedbacks are more commonly found in terrestrial and Earth systems tipping points. The likelihood of cascading effects depends on cross-scale interactions, but differs for each cascading effect type. Regime shifts should not be studied in isolation: instead, methods and data collection should account for potential teleconnections.
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf", dpi = 300)
options(knitr.table.format = "latex")

rm(list = ls())
set.seed(12345)
library(knitr)
library (dplyr)
library (tidyr)
library(forcats)
library (network)
library (sna)
library(RColorBrewer)
library(ggplot2)
library(GGally)
library(viridis)
library(ggrepel)

# for plotting
library(ggmap)
library(maptools)
library(maps)
library(mapproj)

library(gridBase)
library(grid)
library(gridExtra)
library(png)


## for cleaning models
library(car)
library(broom)
library(sandwich); library(lmtest)

### ERGMS for Regime Shifts
library(ergm.count)

## for tables
library(kableExtra)

# setwd('~/Documents/Projects/Cascading Effects/Domino')
# Sys.setenv(RSTUDIO_PANDOC="/Applications/RStudio.app/Contents/MacOS/pandoc")
# rmarkdown::render('170125_draft.Rmd', output_format = 'pdf_document')

  # html_document:
  #   code_folding: hide
  #   dev: pdf
  #   self_contained: yes
  #   toc: yes
  #   toc_float:
  #     collapsed: no
  #     smooth_scroll: no
  # word_document: null

# - \usepackage{lineno}
# - \linenumbers


```


```{r data_ergm, include=FALSE}
# this loads the data, it was cleaned and prepared with script 170329_read_data.R
# and load the ergm models fitted with script 170329_ergms.R

load('~/Documents/Projects/Cascading Effects/170525_ergm_data.RData')


## updated functions ##
## this can be deleted in the future

## Updated ploting function
plotnet <- function(net, ...){
	plot.network(net, #mode='circle',
			vertex.col = alpha(net %v% 'col', 0.7),
			# label = network.vertex.names(net),
			# label.cex= 1,
			main = net %n% 'name', col.main = "gray44", cex.main = 1.2,
			vertex.border = 0,
			usecurve=T,
			vertex.cex= 2, #2 + scale(net %v% 'fb'),
			label.pos= 5,
			edge.col= alpha(net %e% 'col' , 1),
			edge.lwd =  rep(0.01, max(valid.eids(net))), #0.05 + net %e% 'fb',
			edge.curve = 0.01,
			displayisolates=T, pad = 0
			)
}

plotnet2 <- function(net, ...){
	plot.network(net, #mode='circle',
			vertex.col = alpha(net %v% 'col', 0.7),
			# label = network.vertex.names(net),
			# label.cex= 1,
			main = net %n% 'name', col.main = "gray44", cex.main = 1.2,
			vertex.border = 0,
			usecurve=T,
			vertex.cex= 2 + scale(net %v% 'fb'),
			label.pos= 5,
			edge.col= alpha(net %e% 'col' , 1),
			edge.lwd =  0.01 * net %e% 'fb',
			edge.curve = 0.01,
			displayisolates=T, pad = 0 # no need of pad for labels
			)
		}
## A function to create the network, add polarity and cycles / feedbacks as edge attributes
## J161105: Edge attributes are tricky to set because once the network is created the ordering of edges is lost and I cannot find the default order to set the right values. Reading the help of 'loading.attributes' I found that when declared as edgelist one can add attributes directly on the dataframe.
rs.net <- function (dat, i){
	# filter dataset
	dat <- filter(dat, Regime.Shift == levels(dat$Regime.Shift)[i],
	 Polarity == 1 | Polarity == -1)
	#dat <- droplevels(dat)
	# This ordering step is necessary to make sure polarity and other attributes are assigned correctly
	# dat <- dat[order(dat$Tail, partial = dat$Head, decreasing = F),]
	# tab <- table(dat$Tail, dat$Head, dat$Polarity) # tab is an array of tables
		# dimnames(tab) [[3]] "-1"   "-0.5" "0.5"  "1"
		# v <- as.matrix(tab[,,1]*-1) + as.matrix(tab[,,2])
	# build network
		rs.x <- network(select(dat, Tail, Head, Polarity, col), directed = T, ignore.eval=FALSE, matrix.type = 'edgelist')
	# add polarity to edges: Outated, the way I create the network includes already the polarities by using ignore.eval = F. Edge color is also set on the network creation line.
		# rs.x %e% 'polarity' <- dat$Polarity
		# set.edge.value(rs.x, 'polarity', value = as.matrix(v))
		# rs.x %e% 'col' <- ifelse(rs.x %e% "Polarity" == -1, 'red', 'blue')
	# add cycles to nodes and edges.
		fb.sum <- kcycle.census(rs.x, maxlen=network.size(rs.x), mode='digraph',	tabulate.by.vertex=T, cycle.comembership='sum')
		# number of cycles per nodes
		rs.x %v% 'fb' <-  diag(fb.sum$cycle.comemb)
		# number of cycles per edge
		rs.x %e% 'fb' <- as.sociomatrix(rs.x) * fb.sum$cycle.comemb
		#J161106: Note that cycle.comemb is a matrix with comemberships, so two nodes that are not connected in the network can belong to the same cycles. That's why I multiply by the adjacency matrix, to isolate only the edges. cycle.comemb can be used later for discovery of inconvenient feedbacks.
	## add vertex attributes
		rs.x %v% 'col' <- ifelse(colSums(fb.sum$cycle.count)[-1] == 0, "#E41A1C","#8DA0CB") # "#FFD92F"
		rs.x %n% 'name' <- levels(dat$Regime.Shift)[i]
	return(rs.x)
}

```


## Introduction

Regime shifts are large, abrupt and persistent changes in the function and structure of systems [@Scheffer:2003p339; @Scheffer:2001uu]. They have been documented for a wide variety of systems from climate, finance, language, neurological diseases, ecosystems and ancient societies [@Sole:2011us; @Scheffer:2009wl]. In social-ecological systems they present a challenge for ecological management and governance because they may be very difficult to predict [@Boettiger:2013bm; @Hastings:2010p5336] and reverse [@Scheffer:2003p339], while having substantial impacts on the availability of ecosystem services that societies rely upon [@Carpenter:2009jr]. A regime is the region of the parameter space where fast state variables (e.g. vegetation density, coral cover) fluctuate. They may be equilibria, but more generally are basins or domains of attraction. Systems prone to regime shifts have more than one alternative stable state or basin; that is, for the same parameter values (slow variables) they can suddenly shift from one domain to another when critical thresholds are crossed [@MAY:1977p3426; @Holling:1973p6861]. Change in a parameter (e.g. temperature in coral reefs) often shrinks the basin of attraction, making the system more susceptible to shifting when exposed to shock events (e.g. hurricanes) or the action of external drivers [@Scheffer:2012cta].

Over 30 regime shifts in social-ecological systems have been documented [@Biggs:2015iha]; and as human pressures increase on the planet, these shifts might become more acute and frequent. Research on regime shifts is often confined to well-defined branches of science, reflecting empirical, theoretical [@Carpenter2003rsi] or predictive approaches [@Scheffer:2012cta; @Scheffer:2009p4449]. These approaches require a deep knowledge of the causal structure of the of system or a high quality of spatio-temporal data. Hence research on regime shifts has generally focused on the analysis of individual types of regime shifts rather than potential interactions across systems. Yet, one of the key challenges of sustainability science is analyzing the diverse set of interactions across human-environmental systems [@Liu:2015go].

How different regime shifts might be interconnected is largely unknown, and a key frontier of research [@Liu:2015go; @Hughes:2013cv]. Some examples of cascading effects between regime shifts have been reported in the literature, but a systematic review is lacking (See summary Table 1). For example, eutrophication is often reported as a preceding regime shift to hypoxia or dead zones in coastal areas [@Diaz:2008p199]. Similarly, hypoxic events have been reported to affect the resilience of coral reefs to warming and other stressors in the tropics [@Altieri:2017bl]. Moisture recycling is a key underlying feedback on the shift from forest to savanna [@Zemp:2017dr; @Boers:2017ej; @Khanna:2017br] or the Indian monsoon [@Donges:2015fv, @Malik:2011bv]; but also has the potential to couple ecosystems beyond the forest that depend on moisture recycling as an important water source. Changes in moisture recycling can affect mountain forest in the Andes [@Clark:2015bj; @MoruetaHolme:2015fy], nutrient cycling in the ocean by affecting sea surface temperature and therefore regime shifts in marine food webs [@Bakun:2010p5340; @Bakun:2017en], or exacerbation of dry land related regime shifts [@DebraPCPeters:2004ex]. The increasing frequency of boreal forest fire could actually be strong enough to increase warming in the Arctic [@Young:2016kj; @Kelly:2015iq] potentially exacerbating Arctic sea ice loss. By the same token, Arctic sea ice loss can have important impacts on the weakening of the thermohaline circulation and in the long term further exacerbating warming [@Sevellec:2017fk]. Permafrost thawing across the Arctic peatlands can be a dangerous amplifier of global warming too by releasing methane [@Zona:2016hq]. Mangrove forests are expected to store less carbon as their areas shrink and their ability to build peat soils is overwhelmed by sea level rise; however, it is unclear whether the effect will be comparable as to impact global carbon budgets and further warming [@Alongi:2014kq]. While this growing body of literature already presents some potential cascading effects, a systematic assessment is needed of which regime shifts could be interconnected and how.


```{r table_1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}

#
# text_tbl <- data_frame(
#     rs1 = c(" Marine eutrophication", "Hypoxia", rep("Forest to savanna", 3), "Coniferous to deciduos boreal forest", "Peatland transitions", "Arctic sea ice loss"),
#     rs2 = c("Hypoxia", "Coral transitions", "Mountain forest transitions", "Marine foodwebs", "Dryland degradation", "Arctic sea ice loss", "Arctic sea ice loss", "Thermohaline circulation"),
#     type = c("Sharing drivers, domino effect", rep("Domino effect",4), rep("Hidden feedback", 3)),
#     referencia = c("[@Diaz:2008p199]", "[@Altieri:2017bl]", "[@Clark:2015bj; @MoruetaHolme:2015fy]", "[@Bakun:2010p5340]", "[@DebraPCPeters:2004ex]", "[@Young:2016kj; @Kelly:2015iq]", "[@Zona:2016hq]", " [@Sevellec:2017fk]" )
# )
# kable(text_tbl, format = "latex", booktabs = T,
#       col.names = c("Regime shift 1", "Regime shift 2", "Type", "Reference"),
#       caption = "Summary of examples from the literature")

```


: Summary of examples from the literature

| Regime shift 1 | Regime Shift 2 | Type | Reference |
|:--------------:|:--------------:|:--------------:|:--------------:|
| Marine eutrophication | Hypoxia | Sharing drivers, domino effect | [@Diaz:2008p199] |
| Hypoxia | Coral transitions | Domino effect | [@Altieri:2017bl] |
| Forest to savanna | Mountain forest transitions | Domino effect | [@Clark:2015bj; @MoruetaHolme:2015fy] |
| Forest to savanna | Marine food webs | Domino effect | [@Bakun:2010p5340] |
| Forest to savanna | Dryland degradation | Domino effect | [@DebraPCPeters:2004ex] |
| Coniferous to deciduous boreal forest | Arctic sea ice loss | Hidden feedback | [@Young:2016kj; @Kelly:2015iq] |
| Peatland transitions | Arctic sea ice loss | Hidden feedback | [@Zona:2016hq] |
| Arctic sea ice loss | Thermohaline circulation weakening | Hidden feedback | [@Sevellec:2017fk] |


The aim of this paper is to explore potential interactions among regime shifts in social-ecological systems. We propose three types of interconnections: driver sharing, domino effects and hidden feedbacks. Regime shifts are reduced to a network of drivers and feedback processes underlying regime shift dynamics. These directed signed graphs allow us to explore driver co-occurrence, directional pathways and emergent feedback cycles of coupled regime shift networks (See SM Methods). Sharing drivers is likely to increase correlation in time or space among regime shifts, but not necessarily interdependence [@Hughes:2013cv]. Domino effects occur when the feedback processes of one regime shifts affect the drivers of another, creating a one-way dependency [@Rocha:2010vv]. We identified hidden feedbacks by mapping circular pathways on coupled networks that have not been previously reported, helping us detect potential two-way dependencies. The question we are trying to answer whether regime shifts might be interconnected and how - calls for a method for studying patterns of relational data, namely networks. Before outlining the technical details and data used, first we present the hypothesis we test. Then we describe the datasets used and how networks help us identify the three types of interconnections proposed.

## Hypotheses of cascading effects

What the examples from the literature mentioned above (table 1) have in common is that a larger-scale feedback often affects a more localized dynamic in another ecosystem. However, the opposite is also possible, when the accumulation of small-scale processes scale up and impact larger-scale dynamics. The observation that cross-scale interactions can underlie regime-shift interconnections has been previously explored for the dynamics of the dust bowl, desertification, or amplification of fire dynamics[@Anonymous:2007tc; @DebraPCPeters:2004ex]. Therefore our first hypothesis is that cross-scale dynamics underlie potential cascading effects: we expect that domino effects will be mostly dominated by connections between regime shifts that occur in relatively larger scales and slower dynamics in time towards regime shifts in smaller scales and faster in time. The reason to believe so is that a domino effect occurs when a variable that belongs to a feedback mechanism in a regime shift acts as driver in another. Theory often treats drivers as slow variables, which assumes that their change are relatively slower with respect to state variables [@Walker:2012bx]. Conversely, we expect that hidden feedbacks will occur when scales match (both in space and time) given that the pathway between the two regime shifts is bidirectional. We assume that for a new feedback to arise the independent regime shifts need to be somehow aligned in scale for the coupling to work. Additionally, we expect that sharing drivers is more context-specific; in this case, regime shifts occurring in similar ecosystem types or land uses will be subject to relatively similar sets of drivers, increasing the likelihood of driver sharing. This type of connection is expected to increase the likelihood of synchronization in space and time (correlations) of different regime-shift phenomena [@Hughes:2013cv], both in terms of cases (e.g. two different coral reefs) or regime-shift types (e.g. thermokrast lakes and peatland transitions both driven by climate change).

To test these hypotheses we used the [regime-shifts database](www.regimeshifts.org)[@Biggs:2015iha], to our knowledge the largest online repository of regime shifts in social-ecological systems (SM Fig 2, SM Methods). It offers syntheses of over $30$ generic types of regime shifts, $> 300$ case studies based on literature review of $> 1000$ scientific papers [@Biggs:2015iha]. The database decomposes regime shifts in terms of describing their regimes, drivers, feedback mechanisms, impacts on ecosystem services, and management options. It provides a set of $75$ categorical variables (Table 2) about impacts, scales, and evidence types [@Biggs:2015iha] that we used to test our hypotheses. The regime shift database also offers causal-loop diagrams as a summary of the drivers and underlying feedbacks of each regime shift (SM Fig1). Each causal loop diagram was turned into a network by creating the adjacency matrix $A$, where $A_{i,j}$ is $1$ if there is a connection or zero otherwise. A link between two nodes in these networks means that there is at least a scientific paper reviewed in the database providing some evidence for such causal relationship [@Rocha:2015du]. Link sign $w_{i,j}$ represents a link polarity taking $-1$ if the relationship is expected to be negative, or $1$ when positive (See SM). Based on the signed graphs, we merged pairs of regime-shifts networks and created three response variables matrixes that correspond to each of the cascading effects (Fig 1).



```{r table_2, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}
text_tbl <- data_frame(
    Terms = c("Land use", "Ecosystem type", "Ecosystem processes", "Provisioning services", "Regulating services", "Cultural services", "Impacts on human well being", "Spatial scale", "Temporal scale", "Reversibility", "Evidence type"),
    Explanation = c("Land use under which the regime shift occurs", "Ecosystem type in which the regime shift occurs", "Impacts on key ecosystem processes",
                    "Impacts on provisioning ecosystem services", "Impacts on regulating ecosystem services", "Impacts on cultural ecosystem services",
                    "Impacts on aspects of human well being", "Typical spatial scale at which the regime shift occurs",
                    "Typical time scale at which the regime shifts occurs", "Reversibility of the regime shift", "Evidence type in the literature"),
    Variables = c("Urban, small-scale subsistence crop cultivation, large-scale commercial crop cultivation,
                  intensive livestock production (eg feedlots, dairies), extensive livestock production (natural rangelands),
                  timber production, fisheries, mining, conservation,
                  tourism, land use impacts are primarily off-site)",
                  "Marine and coastal, freshwater lakes and rivers, temperate and boreal forests, tropical forests, moist savannas and woodlands, drylands and deserts (below ~500mm rainfall/year), mediterranean shrub (e.g. fynbos), grasslands, tundra, rock and ice, agro-ecosystems, planetary",
                  "Soil formation, primary production, nutrient cycling, water cycling, impacts on biodiversity",
                  "Freshwater, food crops, livestock, fisheries, wild animal and plant products, timber, woodfuel, feed fuel and fiber crops, hydropower",
                  "Air quality regulation, climate regulation, water purification, water regulation, regulation of soil erosion, pest and disease regulation, pollination, natural hazard regulation",
                  "Recreation, aesthetic values, knowledge and educational values, spiritual and religious",
                  "Food and nutrition, health (eg toxins, disease), livelihoods and economic activity, security of housing and infrastructure, aesthetic and recreational values, cultural identity, social conflict, no direct impact",
                  "Local/landscape (e.g. lake, catchment, community), national (country), sub-continental (e.g. southern Africa, Amazon basin), global",
                  "Weeks, months, years, decades, centuries, unknown",
                  "Irreversible (on 100 year time scale), hysteretic (difficult to reverse), readily reversible, unknown",
                  "Models, paleo-observation, contemporary observations, experiments, other")
)

kable(text_tbl, format = "latex", booktabs = TRUE,
      caption = "Summary of categorical variables") %>%
    kable_styling(full_width = TRUE, latex_options = c("scale_down", "HOLD_position"),
                  font_size = 8) %>%
    column_spec(3, width = "4in") #%>%
    # add_footnote(c("Source: the regime shifts database [www.regimeshifts.org]"), notation = "symbol")
```
We tested our hypotheses using as response variable for _drivers sharing_ the number of drivers shared, for _domino effects_ the number of directed pathways that connect two regime shifts, and for _hidden feedbacks_ it is the number of k-cycles (where $k$ denotes cycle length) that emerge on the joined network that do not exist on the separate causal networks (Fig 1, SM Methods). Note that all three response variables are matrices that can be represented as weighted networks in our statistical framework, where nodes are regime shifts and the link depends on the cascading effect described (Fig 1). Thus, our research question can be rephrased as "What is the likelihood of a link between regime shifts in the response variable network, and what features change this likelihood?" As explanatory variables we use the regime-shift database categorical variables ($N=75$, Table 2, see SM methods for processing of variables). We focused on how similar two regime shifts are -regarding the categorical variables they share- and how the similarity increases or not the likelihood of having a link, this is, the likelihood of presenting any of the cascading effects types. All these networks contain weighted links of count data; therefore, the specification for the exponential random graph models follows Krivitsky [-@Krivitsky:2012uo] and a Poisson reference distribution. Using this statistical technique tailored to networks allow us to account for selection bias (e.g. the probability of a node being connected is influenced by the level of connections of its peers) that other linear regression approaches lack.

```{r Fig1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 3, fig.width = 4, fig.align = 'center', fig.cap = "Minimal examples of the cascading effects. For all three types of cascading effects we merged pairs of regime shifts networks and created a response variable matrix that accounted for all pair-wise combination of regime shifts in our sample. For example, for drivers sharing two minimal regime shifts are depicted as causal loop diagrams, drivers are coloured red and variables inside feedbacks purple. The joint network is represented as a 2-mode network with drivers and regime shifts as nodes instead. This representation allow us to study the co-occurrence of drivers across regime shifts, in this minimal example only driver 'a' co-occur in both regime shifts 1 and 2. The response variable matrix counts the drivers shared by all pair-wise combinations of regime shifts. In the example of a domino effect two regime shifts are joined together where driver c in regime shift 2 is part of a feedback process in regime shift 1, creating a one-way dependency (orange colored link) between the two regime shifts. The response variable matrix in this case counts all the one-way causal pathways between pair-wise combinations of regime shifts. The example of a hidden feedback is shown with two minimal regime shifts, each of them with only one driver and one feedback, that when joined together give rise to a new unidentified feedback of length 4 (orange circular pathway). The response variable matrix counts all hidden feedbacks that arise when merging all pair-wise combinations of regime shifts. Labelled matrices of the resulting response variables are available in the supplementary material.", dev = "pdf", dpi = 300, dev.args = list(pointsize = 6)}

# load mini data
mini_dat <- readr::read_csv2("~/Documents/Projects/Cascading Effects/minimal_example_cld_cascading_effects.csv")

# setwd("~/Documents/Projects/Cascading Effects/Figures")

# mini_dat %>% filter (Regime.Shift == 'rs1') %>%
#     select(tail, head, color) %>%
#     # network(directed = T, ignore.eval=FALSE, matrix.type = 'edgelist') %>%
#     # ggnet2(edge.color = "color", arrow.size = 1) # no curved edges, I need them
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = c(0,0,0.5,0.5),
#         vertex.color = alpha(c(rep("red", 2), rep("purple",2)), 0.7 ),
#         vertex.frame.color = NA, vertex.label.color = "grey84",
#         layout = matrix(c(rep(0,4), 4,1,3,2), ncol = 2, nrow = 4),
#         main = "regime shift 1")


# quartz(width = 6, height = 4, pointsize = 7)
# par(mfrow = c(3,4), mar = c(2,2,2,2)) # layout(matrix(1:12,4,3, byrow=T))
# plot.new()
# # grid.newpage() # I need this line to plot on quartz, but not on markdown pdf.
# pushViewport(viewport(layout = grid.layout(3,4)))
#
# ## first row 2 clds
# pushViewport(viewport(layout.pos.row=1, layout.pos.col=1))
# par(fig = gridFIG(), new = TRUE)
# quartz(width = 2, height=2, pointsize = 6)
# mini_dat %>% filter (Regime.Shift == 'rs1') %>%
#     select(tail, head, color) %>%
#     # network(directed = T, ignore.eval=FALSE, matrix.type = 'edgelist') %>%
#     # ggnet2(edge.color = "color", arrow.size = 1) # no curved edges, I need them
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = c(0,0,0.5,0.5), edge.arrow.size = 0.5, edge.width = 0.5, vertex.size = 25, vertex.label.cex = 2,
#         vertex.color = alpha(c(rep("red", 2), rep("purple",2)), 0.7 ),
#         vertex.frame.color = NA, vertex.label.color = "white",
#         layout = layout_with_kk, margin = c(rep(-0.45,4))
#          #matrix(c(rep(1,4), 4,1,3,2), ncol = 2, nrow = 4),
#         )
# quartz.save("fig1a.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# title('a)', adj = 0, cex=5)
# popViewport()
#
# pushViewport(viewport(layout.pos.row=1, layout.pos.col=2))
# par(fig = gridFIG(), new = TRUE)
# # quartz(width = 2, height=2, pointsize = 6)
# mini_dat %>% filter (Regime.Shift == 'rs2') %>%
#     select(tail, head, color) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = c(0,0.5,0.5), edge.arrow.size = 0.5,edge.width = 0.5, vertex.size = 25, vertex.label.cex = 2,
#         vertex.color = alpha(c(rep("red", 1), rep("purple",2)), 0.7 ),
#         vertex.frame.color = NA, vertex.label.color = "white",
#         layout = layout_with_kk , margin = c(rep(-0.45,4))
#         #matrix(c(rep(1,3), 3,2,1), ncol = 2, nrow = 3),
#         )
# quartz.save("fig1b.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()

#
# pushViewport(viewport(layout.pos.row=1, layout.pos.col=3))
# par(fig = gridFIG(), new = TRUE)
# # quartz(width = 2, height=2, pointsize = 6)
# mini_dat %>% filter (RS_joined == 'rs1_2') %>%
#     select(tail, Regime.Shift, color) %>%
#     # network(directed = T, ignore.eval=FALSE, matrix.type = 'edgelist') %>%
#     # ggnet2(edge.color = "color", arrow.size = 1) # no curved edges, I need them
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = FALSE, vertex.shape = c("circle", "circle", "square", "square"),
#         vertex.color = alpha(c(rep("red", 2), rep("blue",2)), 0.7 ), vertex.size = 30, vertex.label.cex = 2,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         edge.width = 0.5, edge.arrow.size = 0.5,
#         layout = matrix(c(1,1,2,2, 2,1,2,1), ncol = 2, nrow = 4),
#         margin = c(rep(-0.45,4))
#     )
# quartz.save("fig1c.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()


m1 <- ggplot(
    data = as.sociomatrix(x, "paths") %>% as_tibble() %>% mutate(tail = rownames(as.sociomatrix(x))) %>%
        gather(key = head, value = paths,  1:30)  %>%
        mutate(tail = as.factor(tail), # %>% forcats::fct_rev(),
               head = as.factor(head) %>% forcats::fct_rev()),
    aes(tail, head)) +
    geom_tile(aes(fill = paths), show.legend = FALSE) + ylab("") + xlab("") +
    scale_fill_gradient2(low = "#FFA50080", mid = "gray84" ,high = "#0000FF80",
                         midpoint = 0, na.value = "grey50",
                         guide = guide_colorbar(title = " ", barheight = 3, barwidth = 0.3)) +
    theme_light (base_size=4) +
    theme(axis.text.x = element_blank(), axis.text.y = element_blank(),
          legend.position = c(0.9,0.5), legend.direction = "vertical")

# pushViewport(viewport(layout.pos.row=1, layout.pos.col=4))
# par(fig = gridFIG(), new = TRUE)
# print(m1, newpage = FALSE)
# popViewport()

## second panel
# pushViewport(viewport(layout.pos.row=2, layout.pos.col=1))
# par(fig = gridFIG(), new = TRUE)
# quartz(width = 2, height=2, pointsize = 6)
# mini_dat %>% filter (Regime.Shift == 'rs3') %>%
#     select(tail, head) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = FALSE,
#         vertex.color = alpha(c("red", rep("purple",3), "red"), 0.7 ), vertex.size = 25,  vertex.label.cex = 2,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         edge.arrow.size = 0.5, edge.width = 0.5, margin = c(rep(-0.45,4))
#         # layout = matrix(c(1,1,0.5,1,1, 4,3,2.5,2,1), ncol = 2, nrow = 5) ,
#         )
# quartz.save("fig1d.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# title ('b)', adj = 0, cex = 5)
# popViewport()

# pushViewport(viewport(layout.pos.row=2, layout.pos.col=2))
# par(fig = gridFIG(), new = TRUE)
# mini_dat %>% filter (Regime.Shift == 'rs4') %>%
#     select(tail, head) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = FALSE,
#         vertex.color = alpha(c("red", rep("purple",4)), 0.7 ), vertex.size = 25,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         edge.width = 0.5, edge.arrow.size = 0.5, vertex.label.cex = 2,
#         margin = c(rep(-0.45,4))
#         #layout = matrix(c(0,0,-1,0,1, 3,2,1,0,1), ncol = 2, nrow = 5) ,
#     )
# quartz.save("fig1e.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()

# pushViewport(viewport(layout.pos.row=2, layout.pos.col=3))
# par(fig = gridFIG(), new = TRUE)
# mini_dat %>% filter (RS_joined == 'rs3_4') %>%
#     select(tail, head, color) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = FALSE, edge.arrow.size = 0.5,
#         vertex.color = alpha(c("red","gray","red", "gray", "red", rep("gray",4)), 0.7 ),
#         edge.width = 0.5,vertex.size = 25, vertex.label.cex = 2,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         margin = c(rep(-0.45,4))
#         #layout = matrix(c(0,0,1,0,0,1,2,3,2,  4,3,2.5,2,1, 2.5, 1, 2.5, 3), ncol = 2, nrow = 9) ,
#     )
# quartz.save("fig1f.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()

m2 <- ggplot(
    data = out %>%
         mutate(Tail = as.factor(Tail), # %>% forcats::fct_rev(),
               Head = as.factor(Head) %>% forcats::fct_rev()),
    aes( Head, Tail)) +
    geom_tile(aes(fill = weight), show.legend = FALSE) + ylab("") + xlab("") +
    scale_fill_gradient(high = "red", low = "gray84", na.value = "grey50",
                         guide = guide_colorbar(title = " ", barheight = 3, barwidth = 0.3)) +
    theme_light(base_size=4) +
    theme(axis.text.x = element_blank(), axis.text.y = element_blank(),
          legend.position = c(0.9,0.5), legend.direction = "vertical")

# pushViewport(viewport(layout.pos.row=2, layout.pos.col=4))
# par(fig = gridFIG(), new = TRUE)
# print(m2, newpage = FALSE)
# popViewport()




## third row

# pushViewport(viewport(layout.pos.row=3, layout.pos.col=1))
# par(fig = gridFIG(), new = TRUE)
# mini_dat %>% filter (Regime.Shift == 'rs5') %>%
#     select(tail, head) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = c(0,0.5,0.5), edge.arrow.size = 0.5,
#                         edge.width = 0.5,
#         vertex.color = alpha(c("red", rep("purple",2)), 0.7 ), vertex.size = 25,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         vertex.label.cex = 2, margin = c(rep(-0.45,4)),
#         layout = layout_with_kk #matrix(c(1,1,1, 3,2,1), ncol = 2, nrow = 3) ,
#         )
# quartz.save("fig1g.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# title ('c)', adj = 0, cex = 5)
# popViewport()
#
# pushViewport(viewport(layout.pos.row=3, layout.pos.col=2))
# par(fig = gridFIG(), new = TRUE)
# mini_dat %>% filter (Regime.Shift == 'rs6') %>%
#     select(tail, head) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = c(0,0.5,0.5), edge.arrow.size = 0.5, edge.width = 0.5, vertex.label.cex = 2,
#         vertex.color = alpha(c("red", rep("purple",2)), 0.7 ), vertex.size = 25,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         layout = layout_with_kk , margin = c(rep(-0.45,4))
#         #matrix(c(1,1,1, 1,2,3), ncol = 2, nrow = 3),
#         )
# quartz.save("fig1h.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()
#
# pushViewport(viewport(layout.pos.row=3, layout.pos.col=3))
# par(fig = gridFIG(), new = TRUE)
# mini_dat %>% filter (RS_joined == 'rs5_6') %>%
#     select(tail, head, color) %>%
#     igraph::graph_from_data_frame(directed = TRUE) %>%
#     igraph::plot.igraph(edge.curved = -0.8, edge.arrow.size = 0.5,
#         vertex.color = alpha(c("red","purple","red", "purple"), 0.7 ),
#         edge.width = 0.5, vertex.size = 25,
#         vertex.frame.color = NA, vertex.label.color = "white",
#         vertex.label.cex = 2, margin = c(rep(-0.45,4))
#         #layout = matrix(c(0,0,1,0,0,1,2,3,2,  4,3,2.5,2,1, 2.5, 1, 2.5, 3), ncol = 2, nrow = 9) ,
#         )
# quartz.save("fig1i.png", type = "png", width = 2, height=2, pointsize = 6, dpi = 300, bg = "white", canvas = "white")
# popViewport()

m3 <- ggplot(data = df_inc,  #%>%
#          mutate(Tail = as.factor(Tail), # %>% forcats::fct_rev(),
#                Head = as.factor(Head) %>% forcats::fct_rev()),
         aes(Tail, Head)) +
    geom_tile(aes(fill = log(inc)), show.legend = FALSE) + ylab("") + xlab("") +
    scale_fill_gradient(low = "#FFA50080", high = "#0000FF80", na.value = "grey50",
                        guide = guide_colorbar(title = "[log]", barheight = 3, barwidth = 0.3)) +
    theme_light(base_size=4) +
    theme(axis.text.x = element_blank(), axis.text.y = element_blank(),
          legend.position = c(0.9,0.5), legend.direction = "vertical")


# pushViewport(viewport(layout.pos.row=3, layout.pos.col=4))
# par(mar=c(2,2,2,2))
# print(m3, newpage = F)
# popViewport()

# detach(package:igraph)

# quartz.save(file="methods_scheme.png", width=7, height=7, pointsize = 7, dpi = 300)

### Attempt with GGally
## First recover the png networks into ggplots objects.
p1 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1a.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p2 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1b.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p3 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1c.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )
p4 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1d.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p5 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1e.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p6 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1f.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )
p7 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1g.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p8 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1h.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )

p9 <- ggplot(data = data_frame(x = 0, y= 0), aes(x,y)) + geom_blank() +
    theme_void() +
    annotation_custom(
        grob = grid::rasterGrob( image = readPNG(
            "~/Documents/Projects/Cascading Effects/Figures/fig1i.png"), interpolate = TRUE),
        xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
        )


plot_list <- list(p1,p2,p3, m1, p4,p5,p6,m2, p7,p8,p9,m3)
pp <- ggmatrix(
    plot_list, ncol = 4, nrow = 3,
    xAxisLabels = c("regime shift 1", "regime shift 2", "joint regime shifts", "response variable"),
    yAxisLabels = c("drivers sharing", "domino effects", "hidden feedbacks"),
    byrow = TRUE, showAxisPlotLabels = FALSE, switch = 'y'
)

pp + theme(strip.text = element_text(size = 6))

```

## Driver sharing

```{r Fig2, cache = TRUE, echo = FALSE, error=FALSE, results = 'hide', warning=FALSE, message=FALSE, fig.height = 2.5, fig.width = 5, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Driver sharing. The distribution of drivers shared per regime shift (a) with respect to the number of drivers each one has (black points) shows that regime shifts in aquatic environments tend to have and share more drivers. The drivers most shared are grouped around variables related to climate change, food production, and urban development (b). WAIS abbreviates West Antarctica Ice Sheet collapse.', dev = "pdf", dpi = 300}

## note that from the ergms script, x is the one mode network of the bipartite network
#x <- mode.1(bip1)[[2]] # x is the regime shifts one-mode network
z <- mode.1(bip1)[[1]]
# and
# bip.edgelist <- bind_rows(bip.edgelist)
# bipmat <- as.matrix(table(bip.edgelist))

df_rs_type <- tibble(
    tail = rsdb$name,
    type = c("Aquatic", "Earth", "Aquatic", "Terrestrial", "Terrestrial", "Aquatic",
             "Terrestrial", "Aquatic", "Aquatic", "Terrestrial", "Aquatic", "Earth",
             "Aquatic", "Aquatic", "Land-water interface", "Aquatic",
             "Aquatic", "Earth", "Land-water interface", "Aquatic","Land-water interface",
             "Land-water interface", "Aquatic","Terrestrial","Terrestrial","Terrestrial",
             "Earth","Land-water interface", "Terrestrial", "Earth")
)

g1 <- ggplot(
    data = as.sociomatrix(x, "paths") %>% as_tibble() %>% mutate(tail = rownames(as.sociomatrix(x))) %>%
        gather(key = head, value = paths,  1:30) %>% left_join(df_rs_type) %>%
        mutate(tail = as.factor(tail) ) %>%
        mutate(tail = forcats::fct_reorder(tail, paths, fun = sum, .desc = FALSE)),
    aes(tail, paths, fill = type, color = type)) +
    geom_boxplot(outlier.alpha = 0.5, alpha=0.5,
                 outlier.size = 0.8) +
    geom_point(
        data = bip.edgelist %>% mutate(value = 1) %>%
            group_by(rs) %>%
            summarize(paths = sum(value)) %>% left_join(df_rs_type, by = c("rs" = "tail")) ,
        aes(rs, paths), color = "black", size = 1.5, show.legend = FALSE
    ) +
    scale_fill_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    scale_color_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    coord_flip() +  xlab("regime shifts") + ylab("number of drivers") + ggtitle("A") +
    theme_minimal(base_size = 6) +
    theme(legend.position = c(0.75, 0.2),
          legend.background = element_rect(fill = "white", linetype = 0),
          legend.text = element_text(size = 4), legend.title = element_text(size = 4))

g2 <- ggplot(
    data = as.sociomatrix(z, "paths") %>% as_tibble() %>% mutate(tail = rownames(as.sociomatrix(z))) %>%
        gather(key = head, value = paths,  1:79) %>%
        group_by(tail) %>% filter(mean(paths) >  1 ) %>% ungroup() %>%
        mutate(tail = as.factor(tail)) %>%
        mutate(tail = forcats::fct_reorder(tail, paths, fun = sum, .desc = FALSE)),
    aes(tail, paths)) +
    geom_boxplot(fill = alpha("grey", 0.7), color = "grey", outlier.alpha = 0.5,
                 outlier.size = 0.8) +
    # geom_point(
    #     data = bip.edgelist %>% mutate(value = 1) %>%
    #         group_by(drivers) %>%
    #         summarize(paths = sum(value)),
    #     aes(drivers, paths), color = "purple"
    # ) +
    coord_flip() +  xlab("most shared drivers\n [mean > 1]") + ylab("number of regime shifts") +  ggtitle("B") +
    theme_minimal(base_size = 6)

m1 <- ggplot(
    data = as.sociomatrix(x, "paths") %>% as_tibble() %>% mutate(tail = rownames(as.sociomatrix(x))) %>%
        gather(key = head, value = paths,  1:30)  %>%
        mutate(tail = as.factor(tail), # %>% forcats::fct_rev(),
               head = as.factor(head) %>% forcats::fct_rev()),
    aes(tail, head)) +
    geom_tile(aes(fill = paths)) + ylab("") + xlab("") + ggtitle("Drivers sharing") +
    scale_fill_gradient(low = "gray84" ,high = "#0000FF80", na.value = "grey50",
                         guide = guide_colorbar("shared drivers", barheight = 0.7)) +
    theme_light (base_size = 6) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5 ),
          legend.position = "bottom")

## Heatmap with ggplot
m2 <- ggplot(
    data = out %>%
         mutate(Tail = as.factor(Tail) %>% forcats::fct_rev(),
               Head = as.factor(Head)), #%>% forcats::fct_rev()),
    aes( Head, Tail)) +
    geom_tile(aes(fill = weight)) + ggtitle("Domino effects") +
    ylab("Independent regime shift") + xlab("Dependent regime shift") +
    scale_fill_gradient(high = "red", low = "gray84", na.value = "grey50",
                         guide = guide_colorbar("domino effects", barheight = 0.7))  +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5 ),
          legend.position = "bottom")

df_domino <- out %>%
        mutate(Tail = as.factor(Tail), Head = as.factor(Head)) %>%
        filter(weight > 0) %>% pull(Tail) %>% fct_count() %>% rename(outdegree = n)

df_domino <- left_join(df_domino,
        out %>%
        mutate(Tail = as.factor(Tail), Head = as.factor(Head)) %>%
        filter(weight > 0) %>% pull(Head) %>% fct_count() %>% rename(indegree = n)
)

df_domino <- left_join(df_domino,
    out %>% group_by(Tail) %>%
        summarize(out_avg = mean(weight), out_max = max(weight)) %>%
        rename(f = Tail)
)

df_domino <- left_join(df_domino,
    out %>% group_by(Head) %>%
        summarize(in_avg = mean(weight), in_max = max(weight)) %>%
        rename(f = Head)
)

df_poly <- data_frame(
    id = rep(factor(c("1", "2")), each = 3),
    x = c(0,0,10, 0,25,25),
    y = c(0,25,25, 0,10,0),
    value = as.factor(c(rep(alpha("orange", 0.4),3), rep(alpha("green", 0.4),3)))
)

g3 <- ggplot(
    data = df_domino %>% left_join (df_rs_type, by = c("f" = "tail")) ,
    aes(y = outdegree, x = indegree)) +
    geom_point( aes(color = type), size = 1.5, show.legend = TRUE) +
    #geom_errorbar(aes(ymin = in_avg, ymax = in_max, color = type)) +
    #geom_errorbarh(aes(xmin = out_avg, xmax = out_max, color = type)) +
    geom_text_repel(data = df_domino %>% filter(out_max > 3 | in_max > 3),
                    aes(label = f), size = 1.5) +
    geom_polygon(data = df_poly,  mapping = aes(x, y, group = id, fill = value), show.legend = FALSE, position = "identity", stat = "identity", inherit.aes = FALSE) +
    scale_fill_manual(guide = "none", values = c(alpha("orange", 0.2),alpha("green", 0.2))) +
    scale_color_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    ggtitle("A") + coord_equal() +
    labs(x = "indegree\n [dependent regime shifts]", y = "outdegree\n [independent regime shifts]") +
    theme_minimal(base_size = 6) +
    theme(legend.position = c(0.8, 0.8),
          legend.background = element_rect(fill = "white", linetype = 0),
          legend.text = element_text(size = 5), legend.title = element_text(size =5)
          )


g4 <- ggplot(
    out %>% filter(weight > 0) %>%
        pull(driv2feed) %>% stringr::str_split(pattern = ", ") %>%
        unlist() %>% as_tibble() %>% rename(driver = value) %>%
        group_by(driver) %>% summarize(n = n()) %>% ungroup() %>%
        mutate(driver = as_factor(driver)) %>%
        mutate(driver = fct_reorder(driver, n, .desc = FALSE)) %>%
        filter(n > 1),
    aes(driver, n)) +
    geom_bar(stat = "identity") + coord_flip() + ylab("number of domino effects") + xlab("key variables")+
    ggtitle('B') + theme_minimal(base_size = 6)



## Heatmap with ggplot
m3 <- ggplot(
    data = df_inc %>%
         mutate(Tail = as.factor(Tail) %>% forcats::fct_rev(),
               Head = as.factor(Head)), #%>% forcats::fct_rev()),
    aes( Head, Tail)) +
    geom_tile(aes(fill = log(inc)))  + ggtitle("Hidden feedbacks") +
    ylab("") + xlab("") +
    scale_fill_gradient(low = "#FFA50080", high = "#0000FF80", na.value = "grey50",
                        guide = guide_colorbar("hidden feedbacks [log]", barheight = 0.7)) +
    theme_minimal(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5 ),
          legend.position = "bottom")  +
    inset(
        grob = ggplotGrob(
            ggplot(data = df_inc, aes(inc)) +
                geom_histogram() + xlab('Hidden feedbacks') +
                theme_light(base_size = 6) ) , xmin = 0, xmax =15 , ymin = 0, ymax = 15)

## I need to make the matrix symmetric so it does count the degrees correctly.
df_inc2 <- data.frame(inc = inc, Head = levels(dat$Regime.Shift)[key[1,]], Tail = levels(dat$Regime.Shift)[key[2,]])

df_inc3 <- bind_rows(df_inc, df_inc2)

g5 <- ggplot(
    data = df_inc3 %>%
        mutate(Tail = as.factor(Tail), Head = as.factor(Head)) %>%
        left_join (df_rs_type, by = c("Tail" = "tail")) %>%
        mutate(Tail = fct_reorder(Tail, inc, fun = mean, .desc = FALSE )) %>% filter(inc > 0)  ,
    aes(Tail, inc, color = type, fill = type)) +
    geom_boxplot( outlier.alpha = 0.5, alpha = 0.5,
                 outlier.size = 0.8, show.legend = TRUE) +
    # geom_point(
    #     data = df_inc3 %>%
    #         group_by(Tail) %>%
    #         summarize(outdegree = sum(inc > 0)),
    #     aes(Tail, outdegree), color = "orange", size = 1.5, shape = 12
    # ) +
    scale_y_log10()+
    scale_fill_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    scale_color_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    coord_flip() +
    xlab("regime shifts") + ylab("hidden feedbacks") + ggtitle("A") +
    theme_minimal(base_size = 6)+
    theme(legend.position = c(0.85, 0.2),
          legend.background = element_rect(fill = "white", linetype = 0),
          legend.text = element_text(size = 4), legend.title = element_text(size = 4)
          )

z <- bind_rows(out_dat)

z <- gather(z,`Inconvenient`,`Expected`, key = Feedbacks, value = count)

z$Feedbacks <- factor(z$Feedbacks)
levels(z$Feedbacks)[2] <- "Hidden"

## J171213: Alternative to g6
 brain <- network(select(dat, Tail, Head, Polarity, col), directed = T, ignore.eval=FALSE, matrix.type = 'edgelist')
 df_brain <- tibble(
     variable = network.vertex.names(brain),
     betweenness = betweenness(brain, cmode = "directed", rescale = TRUE),
     indeg = degree(brain, cmode ="indegree"),
     outdeg = degree(brain, cmode = "outdegree"),
     closeness = closeness(brain, cmode = "suminvdir", rescale = TRUE))

g6 <- ggplot(data = df_brain, aes(betweenness, closeness)) +
    geom_point(aes(color = indeg, fill = outdeg),
               shape = 21, size = 1, stroke = 1, alpha = 0.9) +
    geom_text_repel(data = filter(df_brain, betweenness > 0.015 | closeness > 0.0052),  
                    aes(label = variable), size = 1.5) +
    scale_color_gradient(low = "#6495ED", high = "#00008B", guide = "colorbar", name = " Indegree") +
    scale_fill_gradient(low = "#FFD700", high = "#8B0000", guide = "colorbar", name =  "Outdegree") +
    theme_minimal(base_size = 6) + ggtitle("B") +
    theme(legend.position = "right", legend.direction = "vertical",
          legend.text = element_text(size = 4), legend.title = element_text(size = 4)) #+
    # inset(
    #     grob = ggplotGrob(
    #         ggnet2(brain, size=1, arrow.size = 1.5, arrow.gap = 0.01, edge.alpha = 0.5,
    #                edge.size = 0.1)
    #     ), xmin = 0.015, xmax =0.04 , ymin = 0, ymax = 0.004
    # )


### Insert figures with models here.
## Models for sharing drivers
df1 <- tidy(fit.w1); df1$model <- 'sharing drivers'
df1$term <- c("Non-zero",'Sum', "Landuse","Ecosystem type", "Ecosystem processes", "Provisioning services","Regulating services", "Cultural services", "Impacts on human wellbeing", "Spatial scale: national", "Spatial scale: subcontinental", "Temporal scale: month - year", "Temporal scale: week - month", "Temporal scale: year - decade", "Match spatial scale", "Match temporal scale", "Reversibility", "Evidence type")
## model for domino effects
df2 <- tidy(fit.w2); df2$model <- 'domino effects'
df2$term <- c("Non-zero",'Sum', "Landuse","Ecosystem type", "Ecosystem processes", "Provisioning services","Regulating services", "Cultural services", "Impacts on human wellbeing",
               #"Indegree - Spatial scale: national", "Indegree - Spatial scale: subcontinental", "Indegree - Temporal scale: month - year", "Indegree - Temporal scale: week - month", "Indegree - Temporal scale: year - decade",
               #"Outdegree - Spatial scale: national", "Outdegree - Spatial scale: subcontinental", "Outdegree - Temporal scale: month - year", "Outdegree - Temporal scale: week - month", "Outdegree - Temporal scale: year - decade",
              #"Match spatial scale: local","Match spatial scale: national","Match spatial scale: sub-continental", "Match temporal scale",
              "Spatial scale", "Temporal scale",
              "Reversibility", "Evidence type")
## model for hidden feedbacks
df3 <- tidy(fit.w3a); df3$model <- 'hidden feedbacks'
df3$term <- c("Non-zero",'Sum', "Landuse","Ecosystem type", "Ecosystem processes", "Provisioning services","Regulating services", "Cultural services", "Impacts on human wellbeing",
              "Spatial scale: national", "Spatial scale: subcontinental", "Temporal scale: month - year", "Temporal scale: week - month", "Temporal scale: year - decade", "Match spatial scale", "Match temporal scale",
              "Reversibility", "Evidence type")

df <- full_join(df1, df2) %>% full_join(df3) %>% as_tibble()
# df <-df %>% mutate(estimate = exp(estimate), std.error = exp(std.error)) ## In case you want to read probabilities directly instead of log odds
df <- mutate(df, conf.hi = estimate + std.error, conf.low = estimate - std.error)
df$P <- ifelse(df$p.value <= 0.05, "< 0.05", "> 0.05")
df$term <- as.factor(df$term)

# ## change names of places and reorder levels
df$term <- factor(df$term, levels(df$term)[c(9,16,6,3,2,10,11,1,5, 13,17,14,15,19,18,20,7,8, 12,4)] )
# df$term <- factor(df$term, levels(df$term)[c(17, 28, 11,3,2,23,24,1,5, 26,27, 30,29,31, 6:10,18:22,  12:16, 25,4)] )
df$term <- factor(df$term, rev(levels(df$term)))

df$model <- as.factor(df$model)
df$model <- factor(df$model, levels(df$model)[c(3,1,2)])

## plot results
p <- ggplot(df, aes(estimate, term))+
    geom_point(aes(shape = factor(P)), size = 2, show.legend = T) +
    scale_shape_manual(name = "p value", values = c(19,1)) +
    geom_errorbarh(aes(xmax = conf.hi, xmin = conf.low, height = .25),
                   show.legend = T) + #ggtitle("g)") +
    geom_vline(xintercept = 0, color = 'grey', show.legend = F, linetype = 2)  +
    facet_wrap(~ model, scales = "free_x") + theme_light(base_size = 6) +
    theme(legend.position = "right", plot.margin = margin(t=2,r=5,b=2,l=5,"pt"))


## Combine the figure old way
# gg <- list (g1,g2,g3,g4,g5,g6, p)
# source('~/Dropbox/Code/multiplot.R')
# layout <- matrix(c(1:6,7,7,7), ncol = 3, nrow = 3, byrow = T)
# multiplot(plotlist = gg, layout = layout)


## New way:
gg <- list (g1,g2)
source('~/Dropbox/Code/multiplot.R')
layout <- matrix(c(1,2), ncol = 2, nrow = 1, byrow = T)
multiplot(plotlist = gg, layout = layout)

```

Aquatic regime shifts tend to have and share more drivers, although the driver sharing is not exclusively with other aquatic regime shifts (Fig 2a, SM Fig 3). The resulting matrix for _driver sharing_ is a one-mode projection of the 2-mode network composed by `r dim(bipmat)[1]` drivers and `r dim(bipmat)[2]` regime shifts (Fig 1, SM Fig1). The highest concentration of driver sharing is found between regime shifts in kelps, marine euthrophication, the collapse of fisheries and hypoxia. Terrestrial and polar regime shifts tend to have fewer and more idiosyncratic sets of drivers. Large-scale regime shifts in polar and sub-continental areas (e.g. Monsoon weakening) have fewer drivers but are hotspots of sharing, typically including climate-related drivers. In support of previous results [@Rocha:2015du], we find that the most co-occurring drivers are related to food production, climate change and urbanization (Fig 2b). Regime shifts are more likely to share drivers when they occur in similar ecosystem types and impact similar regulating services (p << 0.001, Fig 3, SM Table 1). The odds of two regime shifts sharing drivers is `r round(exp(sum(fit.w1$coef[1])),2)`, while the odds of them sharing drivers given that they occur in similar ecosystem type is `r round(exp(sum(fit.w1$coef[c(1,4)])),2)`. The likelihood of sharing drivers is also affected by occurring on similar land use, impacting similar cultural services and aspects of human wellbeing, matching spatial scales (but not temporal ones), as well as having similar types of evidence and reversibility (p < 0.05, Fig 3, see SM Table 1 for comparison with null and alternative models fitted).


```{r Fig3, cache = TRUE, echo = FALSE, error=FALSE, results = 'hide', warning=FALSE, message=FALSE, fig.height = 2, fig.width = 4, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Statistical models summary. The response variables are the matrices of regime shifts interconnections for each cascading effect type. The terms fitted correspond to the categorical variables from the regime-shifts database (Table 2) that capture the similarity of regime shifts regarding their impacts on ecosystem services and human well being, the scales at which they occur, reversibility and evidence type (See SM Methods). Results from exponential random graph models are interpreted similarly to those of logistic regression. Only best models are shown (higher AIC or BIC), for complete results of null and alternative models fitted see Tables SM1, SM2 and SM3.', dev = "pdf", dpi = 300}

p

```

Intuitively, the sharing of drivers is proposed as a potential mechanism that can correlate regime shifts in space and time, but not necessarily make them interdependent unless they also share feedback mechanisms. Only 35% of all pair-wise combinations of regime shifts are coupled solely by driver sharing. Time correlation is debated given that spatial heterogeneity can break the synchrony induced by the sharing of drivers [@Hughes:2013cv], meaning that contextual settings matter for such correlations to emerge. Spatial heterogeneity is also attributed as a mechanism that can smooth out critical transitions and soften their abruptness [@Martin:2015kl; @Medeiros:2016vf]. Yet, with or without masking mechanisms, identifying common drivers is useful for designing management strategies that target bundles of drivers instead of well-studied variables independently, increasing the chances that managers will avoid several regime shifts under the influence of the same sets of drivers [@Rocha:2015eea; @Rocha:2015du]. For example, management options for drivers such as sedimentation, nutrient leakage, and fishing can reduce the likelihood of regime shifts such as eutrophication and hypoxia in coastal brackish lagoons, as well as coral transitions in adjacent coral reefs.

## Domino effects

```{r Fig4, cache = TRUE, echo = FALSE, error=FALSE, results = 'hide', warning=FALSE, message=FALSE, fig.height = 2, fig.width = 5, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Domino effects. Regime shifts that produce most domino effects (high outdegree, shaded green area) are Earth system tipping points, while the regime shifts that receive the most (high indegree, orange shaded area) occur in aquatic and land-water interface (a); labels are plotted only for regime shifts where the maximum number of domino effects is found (4). Most variables associated with domino effects are related to climate and transport mechanisms (b). These variables are part of a feedback mechanism in one regime shift that are in turn drivers in another regime shift.', dev = "pdf", dpi = 300}

## New way:
gg <- list (g3,g4)
source('~/Dropbox/Code/multiplot.R')
layout <- matrix(c(1,2), ncol = 2, nrow = 1, byrow = T)
multiplot(plotlist = gg, layout = layout)


```

Domino effects were investigated by searching variables that belong to feedback mechanisms in one regime shift and at the same time are drivers of another (Fig 1, see SM Methods). Despite having relatively fewer drivers, regime shifts that account for most domino effects include the Greenland ice sheet collapse, the Monsoon weakening, primary production in the Arctic Ocean, river-channel changes, soil salinization, weakening of the thermohaline circulation and the shift from tundra to forests (Fig 4). Conversely, the regime shifts that receive most influence through domino effects are mangrove transitions, kelp transitions and transitions from salt marshes to tidal flats. The maximum number of pathways found was $4$, and the variables that produce most domino effects relate to climate, nutrients and water transport (Fig 4). In line with our expectations, regime shifts that contain variables that will in turn be drivers of other regime shifts typically have large spatial scales and slow temporal scales: thermohaline circulation collapse, river channel change, monsoon weakening, and Greenland ice sheet collapse. On the other hand, regime shifts that receive the influence are often marine and their time and space dynamics contained more locally. The statistical models support this observation (Fig 3): regime shifts whose time scales are on the range of weeks to months are more likely to receive influence from regime shifts whose dynamics occur on the scale of years to decades (p < 0.1), but we did not find evidence for spatial scales (Table SM2). Both, having a link and having a high number of pathways are significant (p < 0.01). The odds of having higher numbers of domino effects are increased when regime shifts impact similar regulating services and similar aspects of human well-being (p < 0.001).

## Hidden feedbacks

```{r Fig5, cache = TRUE, echo = FALSE, error=FALSE, results = 'hide', warning=FALSE, message=FALSE, fig.height = 2.5, fig.width = 6, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Hidden feedbacks. Hidden feedbacks occur typically in terrestrial and Earth system regime shifts, (a) shows the distributions of hidden feedbacks and has been organised by higher to lower mean in the number of feedbacks. Boxplots are shown in log-scale after zero values have been removed. The variables most often involved in hidden feedbacks have high betweenness and closeness centralities (b) calculated on the network of all regime shifts in our sample (N = 30). These measures reveal the variables (labelled) that lie on most shorter pathways from all other variables in the network', dev = "pdf", dpi = 300}

## New way:
gg <- list (g5,g6)
source('~/Dropbox/Code/multiplot.R')
layout <- matrix(c(1,2), ncol = 2, nrow = 1, byrow = T)
multiplot(plotlist = gg, layout = layout)

```

Hidden feedbacks are feedback mechanisms that might connect two different regime shifts; and if strong enough, they could amplify or dampen the coupled dynamics (Fig 1). Contrary to the results found for driver sharing, most hidden feedbacks occur in terrestrial and earth systems (Fig 5) and typically have higher feedback length. The regime shifts with higher numbers of connections (16 to 18 out of 30 possible) are forest to savanna, monsoon weakening, thermohaline circulation, desertification, primary productivity of the Arctic Ocean and the Greenland ice sheet collapse. Key variables that belong to many of these hidden feedbacks are related to climate, fires, erosion, the functional role of herbivores, agriculture and urbanisation (Fig 5). Even for small networks, the number of cycles tend to increase exponentially with respect to the increase of links. The statistical analysis (Fig 3) shows that there is fewer hidden feedbacks that one would expect by random, but when they do occur, the odds of having multiple feedbacks coupling two regime shifts is `r round(exp(1.99143),2)` times higher. The odds of two regime shifts been connected through hidden feedbacks is not affected by occurring on similar land uses or ecosystem types. Yet, the likelihood of hidden feedbacks increases if the pair of regime shifts impact similar ecosystem processes, impact similar regulating and cultural services, and especially if they match spatial scales but not necessarily temporal ones (Fig 3, p < 0.001, see SM Table 3 for null and alternative models fitted), suggesting cross-scale interactions in time.

Out of the 870 pair-wise combinations of regime shifts analysed (when taking into account directionality), ~35% are solely coupled through sharing drivers, ~4% through domino effects, and ~2% through hidden feedbacks (Fig 6). 28% of these pair-wise combinations are coupled through multiple (2 different types) of cascading effects, and 10% by the three of them. These results highlights that, at least structurally, cascading effects are plausible in $~80$% of all combinations of regime shifts. Only for 167 pair-wise combinations we can be certain with the current dataset that there is no cascading effects. However, the discovery of new drivers or feedback mechanisms underlying these dynamics can reduce this estimate.

```{r Fig6, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 4, fig.width = 7, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Summary of cascading effects. Networks of regime shifts interactions are shown by cascading effect type. These maps show regime shifts in approximately an area of the world where they are expected to occur. Regime shifts are linked if a driver-sharing, domino effect, or hidden feedback was found in our analysis. Only exclusive connections are shown, figure SM2 complements this figure by showing the matrix where multiple connections are expected as well as the pair-wise combinations where no connections were found. Here we also show the network where all three types of cascading effects are expected to occur.', dev = "pdf", dpi = 300}

df_fork <- as.sociomatrix(x, "paths") %>% as_tibble() %>% mutate(tail = rownames(as.sociomatrix(x))) %>%
        gather(key = head, value = drivers,  1:30) %>% mutate(tail = as.factor(tail) %>% forcats::fct_rev()) %>%
    mutate(drivers = ifelse(drivers > 0, 1, 0))
# domino <- as.sociomatrix(dom_net)
# inconvenient <- as.sociomatrix(inc_net)

# inconvenient <- cbind(inconvenient, rep(0,29))
# inconvenient <- rbind(inconvenient, rep(0,30))
#
# colnames(inconvenient)[30] <- "Sprawling vs compact city"
# rownames(inconvenient)[30] <- "Sprawling vs compact city"
#
# inconvenient <- clean.and.order(inconvenient)

# all <- as.data.frame(fork + domino + inconvenient)
# all$from <- rownames(all)
# df_all <- gather(all,  1:30, key = to, value = count)


df_domino <- as_tibble(out) %>% rename(tail = Tail, head = Head) %>%
    mutate(domino = ifelse(weight > 0, 1, 0)) %>% select(-driv2feed, -weight)
df_hidden <- as_tibble(df_inc3) %>% rename(tail = Tail, head = Head) %>%
    mutate(hidden = ifelse(inc > 0, 1, 0)) %>% select(-inc)


df_all <- left_join(df_fork, df_domino) %>% left_join(df_hidden) %>%
    group_by(tail,head) %>%
    mutate(all = sum(drivers, domino, hidden),
           type = ifelse(is.na(all), NA,
                         ifelse(all == 3, "all",
                                ifelse(all == 2, "multiple",
                                       ifelse(all == 1 && drivers == 1, "drivers sharing",
                                              ifelse(all == 1 && domino == 1, "domino effect",
                                                     ifelse(all == 1 && hidden == 1, "hidden feedback", NA)
                                                     )
                                              )
                                       )
                                )
           )
    )

df_all$type <- as.factor(df_all$type) %>% forcats::fct_relevel(
    "drivers sharing", "domino effect", "hidden feedback", "multiple", "all") %>%
    forcats::fct_rev()


### Experiment with maps as summarising figure.
## Here is the code copied from chunk below to create the world maps: re do the coordinates only with RS used in the paper.
place <- c(
    "Svalbard", # "Arctic Benthos Borealisation"
    "North pole",#"Arctic Sea-Ice Loss"
    "Chesapeake bay", #"Bivalves collapse"
    "Serengeti National Park",# "Bush encroachment"
    "Alaska",#"Coniferous to deciduous forest"
    "Great Barrier Reef",#"Coral transitions"
    "Sahara desert", # "Desertification"
    "Southern Ocean",#"Fisheries collapse"
    "Wolderwijd",#"Floating plants"
    "Amazon, Brazil",#"Forest to savanna"
    "laguna de Fuquene",#"Freshwater eutrophication"
    "Greenland",#"Greenland Ice Sheet collapse"
    "Gulf of Mexico",# "Hypoxia"
    "Farallon Islands",#"Kelps transitions"
    "Thailand",#"Mangroves transitions"
    "Baltic Sea",# "Marine eutrophication"
    "Benguela, Namibia",#"Marine foodwebs"
    "India",#"Moonson"
    "Finland",# "Peatland transitions"
    "Arctic ocean",#"Primary production Arctic Ocean"
    "Lena river, Russia",#"River channel change"
    "Elizabeth River, Portsmouth, VA",# "Salt marshes to tidal flats"
    "Turks and Caicos Islands",#"Seagrass transitions"
    "Australia",#"Soil Salinization"
    "Los Angeles",# "Sprawling vs compact city"
    "Pleistocene Park, Sakha Republic, Russia",#"Steppe to Tundra"
    "Iceland",#"Thermohaline circulation"
   "Russkoye Ustye, Russia", # "Thermokarst lakes"
    "Yukon",#"Tundra to forest"
    "West Antartica" #"WAIS"
              )
#
# rs_coords <- tibble(name = rsdb$name, place)
# coord <- geocode(place, source = "google")
#
# ## repeat until complete!
# coord[is.na(coord$lon),] <- geocode(place[is.na(coord$lon)], source = "google")
#
# rs_coords <- rs_coords %>% mutate(
#     x = coord$lon,
#     y = coord$lat
# )


# write.csv(rs_coords, file = 'rs_coords_paper.csv' )

rs_coords <- read.csv(file = '~/Documents/Projects/Cascading Effects/Domino/rs_coords_paper.csv', dec = '.')

## correct some manually
rs_coords[rs_coords$name == "Arctic Sea-Ice Loss", 3:4] <- c(0, 80)
rs_coords[rs_coords$name == "Fisheries collapse", 3:4] <- c(-80, -67)
rs_coords[rs_coords$name == "Tundra to forest", 3:4] <- c(-110, 60)
rs_coords[rs_coords$name == "Salt marshes to tidal flats", 3:4] <- c(-76.34355, 26.92487)


world <- map_data('world')
# world <- fortify(world)

m <- ggplot(world, aes(long, lat)) +
    geom_polygon(fill="grey90", aes(group = group)) +
    geom_path(color="white",aes(group=group), size = 0.2) +
    coord_equal() + theme_void(base_size = 7, base_family = "Helvetica")

## Once the map is created I need to merge the coordinates and the data frame for all cascading effects to create the networks:

df_all <- df_all %>% left_join(
    select(rs_coords, -place), by = c("head" = "name")) %>%
    rename(xend = x, yend = y) %>%
    left_join(select(rs_coords,-place), by = c("tail" = "name"))

mp_all <- m + geom_segment(data = df_all %>% filter(!is.na(type), type != "multiple"),
                 aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.2, size = 0.2) +
    # geom_point(data = rs_coords, aes(x,y)) +
    geom_text(data = left_join(rs_coords, df_rs_type, by = c("name" = "tail") ) %>%
                  rename(rstype = type),
              aes(x,y, label = name, color = rstype), size = 1) +
    scale_color_manual("Ecosystem type",values = c("blue", "purple", "red","orange")) +
    theme_void(base_size = 7) + facet_wrap(~type) +
    theme(legend.position = "bottom")

mp_all

```

## Discussion

We have identified structural dependencies that can give rise to cascading effects among different regime shifts. Of particular importance, the weakening of the Indian Monsoon, the thermohaline circulation, desertification, forest-to-savanna or seagrass transitions are regime shifts that are expected to be involved in many of these couplings (Fig 6). We hypothesised that the likelihood of sharing drivers increased when regime shifts occur in similar ecosystem types and land uses. We confirmed this observation and also found that sharing drivers increase when regime shifts occur on the same spatial scales, or when they impact similar provisioning or regulating services. Based on examples of cascading effects already reported in the literature (Table 1) we hypothesised that domino effects will be more common between regime shifts whose dynamics occur at larger spatial scales and slower dynamics in time towards regime shifts more localised in space and faster in time. We only found support for the cross-scale dynamic in time but not in space. Additionally, regime shifts that often received the domino effect tend to occur in aquatic environments pointing out the importance of water transport as a potential mechanism linking regime shifts in one-way interactions. Last, we hypothesised that hidden feedbacks would appear when scales match both in space and time, as well as under similar ecosystem types. Our results only confirm matching in spatial scales and also suggest that hidden feedbacks are common when regime shifts impact similar ecosystem processes, regulating and cultural ecosystem services.

Domino effects and hidden feedbacks are often disregarded because research on regime shifts usually focus on one system at the time; data collection and hypothesis testing for coupled systems has largely remained unexplored. In fact, the question of how different regime shifts might be interconnected remains a key frontier of research [@Liu:2015go; @Hughes:2013cv]. Our results show that $~45$% of all pair-wise combinations of regime shifts imply a plausible domino effect or hidden feedback (Fig 6). Recent literature reports potential linkages between euthrophication and hypoxia [@Diaz:2008p199], hypoxia and coral transitions [@Altieri:2017bl], or shifts in coral reefs and mangroves transitions [@Alongi:2014kq]. Other examples in the terrestrial realm report potential increase in Arctic warming from higher fire frequency in boreal forest [@Young:2016kj; @Kelly:2015iq] or permafrost thawing [@Zona:2016hq], potentially impacting any regime shift in the Arctic driven by temperature [@Peterson:2016ul]. Shifts in forest that can reduce moisture recycling has also been reported to potentially impact mountain forest [@Clark:2015bj; @MoruetaHolme:2015fy], drylands [@DebraPCPeters:2004ex], and regime shifts in the ocean affected by nutrient cycling [@Bakun:2010p5340]. While this handful of examples represent an emergent literature on regime shifts interactions, the method here developed allow us to explore more rigorously the type of potential couplings between regime shifts. Our findings align with previous results on the type of variables and processes that might couple distant regime shifts, highlighting the role of climate and transport mechanisms for nutrients and water (Fig 2,4,5). An important lesson from our study is that regime shifts can be interconnected, they should not be study in isolation assuming they are independent systems. On the contrary, methods and data collection that takes into account the possibility of cascading effects needs to be further developed.

Here we have advanced a network-based method to explore plausible cascading effects and distinguish potential correlations from true interdependencies. For our sample of 30 regime shifts reported in the regime shifts database, 435 possible pair-wise combinations exist -870 with directionality. Exploring such search space is not a trivial task. Once the mechanisms responsible for one coupling are identified, it still remains to explore the parameter space under which the coupling is feasible. In other words, while our method identifies plausible connections, modelling efforts and observational studies are imperative to distinguish plausible from probable. Dynamic models of this type of dynamics require careful assumptions about parameter values as well as functional form of the system's equations. Generalised modelling is a promising technique that does not require particular assumptions allowing the researcher to reach more general conclusions based on stability properties of the system [@Gross:2009jr; @Gross:2004uu; @Lade:2013iwa]. Another potential avenue for future research is looking at how transport mechanisms couple far-apart ecosystems. One example already mentioned is the moisture recycling feedback and how it can affect the water budget of areas down the '_preceipitationshed_' [@Keys:2012bo]. Another teleconnection (sensu [@Liu:2015go]) could be with allocation of resources through international trade, investigating how demand of resources in certain countries can shape the state space of ecosystems from the providing countries. While the literature continues to provide examples of potential couplings, our results contribute to the ongoing debate by distinguishing whether the coupling is expected to be correlational because of driver sharing, a one-way causation (the _domino effect_), or a two-way interaction (the _hidden feedbacks_), providing a useful set of hypotheses for future research.

The fact that regime shifts can be interconnected raises management and governance challenges in the Anthropocene. It means that what happens in a regional ecosystem can be influenced by ecosystem management far away, and conversely, can have implications for shifts in other ecosystems. Decisions in one place can undermine the achievement of sustainable development goals in other places. For example, it has been shown that what happens in the Arctic does not stay in the Arctic [@Cvijanovic:2017hn; @Peterson:2016ul], many Arctic regime shifts have the potential to impact non-Arctic ecosystems far away and the provision of their ecosystem services. It implies that whoever does make decisions on management is not necessarily the one that has to deal with the impacts. This issue is evident in governance of water transport systems, whether run-off or atmospheric transport, but it is applicable to other dynamics connecting far-away ecosystems through other mechanisms such as climate change, fire, nutrient inputs or trade. Our results highlight variables that are key for domino effects and hidden feedbacks. They are also good observables for monitoring early warning indicators of the strengthening of regime-shifts coupling.

## Conclusions

Non-linear dynamics are ubiquitous in a large range of social-ecological systems. However, how a regime shift somewhere in the world could affect the occurrence of another regime shift remains an open question and a key frontier of research. Here we have developed a network-based method to explore potential cascading effects among regime shifts. Based on topological features of causal networks, we have proposed three type of connections. Potential correlations can arise when regime shifts share common drivers. While the spatio-temporal correlation can be masked by heterogeneity or noise, the cluster of drivers shared serves to design managerial strategies. One-way directional interactions are denoted as _domino effects_ and are common but not exclusive to aquatic regime shifts. These connections tend to emerge between regime shifts whose dynamics occur at similar temporal scales. Two-way interconnections are denoted _hidden feedbacks_ and occur often between regime shifts that match spatial scales and impact similar ecosystem processes. The two latter types of cascading effects call for a more holistic approach for modelling and studying regime shifts, acknowledging their potential interdependencies.

\pagebreak


## Supplementary material

This section complements the methods and results from the exponential random graph models. The modeling technique used the number of drivers shared, the number of domino effects and the number of hidden feedback as the three key response variables (Fig 1). The use of exponential random graph models is motivated because it take into account network structure and allow us to control for sample bias. It investigates the odds of the existence and weight of a link and its results are interpreted similarly to a logistic regression.


### Supplementary methods:

_Data_: The [regime shifts database](www.regimeshifts.org) synthesizes scientific literature on our current knowledge on regime shifts. It is an open online repository of regime shift syntheses and case studies. When possible, entries to the regime shifts database have been peer-reviewed by an expert on the topic to ensure quality and accuracy of its contents [@Biggs:2015iha]. The main data used in this study are the causal loop diagrams described in the database (SM Fig 1), and a set of categorical variables about regime shifts attributes that were used to fit statistical models to explore the role of cross scale interactions (described below). Out of the 35 regime shifts documented in the database, here we use 30 where complete synthesis exist and causal loop diagrams have been curated.

_Causal loop diagrams_ (CLDs) represent a collection of causal mechanisms that scientist have reported in their narratives: both empirical (what they choose to sample) or theoretical (what they choose to model). CLDs consist of variables connected by arrows denoting causal influence [@Sterman:2000we], also known as signed directed graphs. Each relationship must have a positive $(+)$ or negative $(-)$ sign that represents the effect of the dependent variable given change on the independent variable [@Lane:2008p6781; @Sterman:2000we]. Although the functional form that underlies each relationship is not necessarily known, positive relationships are proportional while negative ones are inversely proportional _ceteris paribus_. CLDs assume that the causal relationships captured by links are monotonic, while non-linearities are captured by links sets: thus they do not have self-loops. Feedback loops are the basic structural units of the diagram and emerge by connecting variables in closed directed paths (cycles). Feedback means that once a signal enters the loop, some part of the output is fed back to the input, resulting in amplification or dampening of its own signal. Feedbacks can be reinforcing if the overall polarity of its links is positive, or balancing if negative. Reinforcing feedbacks are usually responsible for behaviors that drive the system out of equilibrium, while balancing feedbacks are responsible for near equilibrium dynamics such as oscillations and delays [@Sterman:2000we]. Note that causal links do not describe the behavior of variables, only the structure of the system: they describe what would happen if there were changes [@Sterman:2000we]. CLDs were curated in the regime shifts database in a way that variables names are consistent (e.g. agriculture and cropping is kept as 'agriculture'), and feedback loops comparable (e.g. albedo in the rainforest, Arctic or Antarctic regime shifts is the same feedback). SM Fig 1 shows all CLD's used in the analysis.


```{r supp1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 10, fig.width = 10, dev.args = list(pointsize= 7), fig.align='center', fig.cap = 'Supplementary figure 1. Causal networks for all regime shifts analysed'}

## Updated ploting function

# detach(package:igraph)
library(network); library(sna)
# quartz(pointsize = 7)
par(mfrow = c(6,5), mar = c(2,2,2,2))

for (i in 1:length(levels(dat$Regime.Shift)) ){
	net <- rs.net(dat = dat,  i = i)
	plotnet(net)

}
```

_Categorical variables:_ We calculated the similarity of each pair-wise combination of regime shifts in the database regarding categorical attributes (Table 1 main text) such as *(i)* land use under which the regime shift occur, *(ii)* ecosystem type, impacts on *(iii)* ecosystem processes, *(iv)* provisioning services, *(v)* regulating services, *(vi)* cultural services, *(vii)* and human wellbeing; as well as *(viii)* the spatial scale at which the regime shift occur, *(ix)* the temporal scales, *(x)* reversibility, and *(xi)* evidence type [@Biggs:2015iha; @Rocha:2015du]. For all $75$ categorical variables encoded, the database reports presence or absence $(0,1)$ allowing us to calculate the Jaccard index and use it as a proxy of how similar two regime shifts are. To facilitate the interpretation of statistical models, the Jaccard distance was rescaled ($x =1 - J_{d}$) so equivalent regime shifts score 1 and complete dissimilar zero. Given that most of our hypotheses relate to the scale at which regime shifts occur, we also modelled scale as a categorical variable to account for matches in the network, not only similarity. SM Fig 2 shows a map with all generic types of regime shifts and case studies reported in the regime shifts database.


```{r SuppFig1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 4, fig.width = 6, fig.align = 'center', fig.cap = "Regime shifts around the world. Large points show generic types of regime shifts (n = 35) while small points are case studies (n=324).", dev.args = list(pointsize = 5)}

coords <- read.csv2("~/Documents/Projects/Cascading Effects/Domino/case_coords.csv", dec = '.')

coord_rs <- read.csv(file = '~/Documents/Projects/Cascading Effects/Domino/rs_coords.csv', dec = '.')
cases <- cbind(cases, coords)
rs <- cbind(rs, coord_rs)
rs$rs <- as.factor(rs$rs)
cases$Type.of.regime.shift <- as.factor(cases$Type.of.regime.shift)
names(cases)[7] <- 'rs'

# correction on RS names for plotting
levels(rs$rs) [17] <- "Marine food webs"
levels(rs$rs) [28] <- "Thermokarst lakes"
levels(cases$rs)[17] <- "Fisheries collapse"
levels(cases$rs)[c(5,22)] <- "Marine food webs"
levels(cases$rs)[c(2,9,12,17,20)] <- "Other"


world <- map_data('world')
world <- fortify(world)

m <- ggplot(world, aes(long, lat)) +
    geom_polygon(fill="grey90", aes(group = group)) +
    geom_path(color="white",aes(group=group), size = 0.2) +
    coord_equal() + theme_void(base_size = 7, base_family = "Helvetica")

m + geom_point(data = cases, aes(lon, lat, color = rs), size = 0.2, show.legend = F) +
    geom_point(data = rs, aes(lon,lat, color = rs), size = 2, show.legend = T ) +
    theme(legend.position = "bottom", legend.text = element_text(size = 4)) +
    scale_color_discrete(guide = guide_legend(title = "Regime shifts documented on the regime shifts database \n Source: www.regimeshifts.org", title.position = 'top', title.hjust = 0.5))

# png(file="rsdb_map.png", width=1280, height = 768, units = "px")
# dev.off()
# quartz.save("rsdb_map.png", width= 12, height=7, pointsize = 10, dpi = 800, bg = "white")

```

_Networks:_ Causal loop diagrams were reinterpreted as networks where node attributes were coded if a node belonged to a feedback loop -a k-cycle in the network- or not. The later is then by definition a driver, an independent variable whose dynamics are not affected by the dynamics of the state variables of the system at hand. Note that networks in ecology usually describe inter-species interactions such as predation or mutualism. Our approach is different from what has been done in ecology. Here a network describes a set of processes, both biotic and abiotic, that can govern ecosystem regime-shift dynamics. Our approach is inspired by other network applications to processes such as cell metabolic networks or the network of human diseases, where a process is not captured by a link type (e.g. predation) but by a collection of link interactions (e.g. the Krebs cycle). Thus, individual species are not taken into consideration, but rather their functional roles at the aggregated ecosystem scale (e.g. herbivory).

_Shared drivers:_ To study shared drivers we created a bipartite network of drivers and regime shifts following the method outlined by Rocha et al [@Rocha:2015du]. The statistical models were performed for the case of driver sharing on the one-mode network projection of regime shifts sharing drivers (this is the matrix $A^TA$), and categorical variables from the regime-shifts database were used as node attributes or node covariates to test our hypothesis.

_Domino effects:_ occur when the occurrence of a regime shift can increase or decrease the likelihood of other regime shifts occurring, creating a one-way dependence. Different from the driver-sharing network, here we explore potential domino effects by using the full causal loop diagrams as networks. The algorithm for identifying domino effects takes the adjacency matrix of two given regime shifts $A_1$ and $A_2$ and identifies all nodes $n \in A_1 \cap A_2$ such that $n$ belong to a feedback in $A_1$ but is a driver in $A_2$. Thus, set differences between causal pathways suggest missing drivers, and set intersection between causal pathways and feedback loop nodes indicate potential domino effects (Fig. 2b). By iterating this simple algorithm we derived how many different pathways exist between every pair-wise combination of regime shifts $(N=870)$. The resulting non-symmetrical matrix represents a directed network with regime shifts as nodes and link weights as the number of pathways used for statistical analysis.

_Hidden feedbacks:_ We explore hidden feedbacks by pair-wise comparison of causal networks. First, the feedback loops (or k-cycles) are counted by feedback length $k$ for each regime shift matrix  $A_1$ and $A_2$ separately. Then the cycle count is applied to the composite network $A_{1,2}$ of the two regime shifts. The difference between the k-cycles in the composite network $A_{1,2}$ and the k-cycles in the individual networks $A_1$ and $A_2$ are the hidden feedbacks that emerge when the two causal networks are joined. By iterating the same procedure to all pair-wise combinations of regime shifts $(N=435)$ we obtain a symmetric directed matrix that is then used for the statistical analysis. Although computationally intensive, the search for cycles is feasible in our networks given the small size and relative sparse structures. In fact, the maximum cycle length $k$ is bounded by the size of the network. Out of the `r length(out_inc)` coupled networks analysed, the maximum feedback length was `r unlist(lapply(out_inc, network.size) ) %>% max`.

```{r matrixes_fig, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 7, dev.args = list(pointsize= 6), fig.align='left', fig.cap = 'Response variable matrices and summary of all cascading effects in matrix form.', dev = "pdf", dpi = 300 }

m4 <- df_all %>% ungroup() %>% # filter(fork == 1, domino ==0, inc ==0) %>%
    mutate(tail = as.factor(tail) %>% forcats::fct_rev(),
               head = as.factor(head)) %>% #%>% forcats::fct_rev()),
    ggplot(aes(y = tail, x = head)) +
    geom_tile(aes(fill= type)) +
    scale_fill_brewer(type = "seq", palette = "YlOrBr", direction = -1, na.value = "gray57",
                      guide = guide_legend(title = "Cascading effects", barwidth = 0.5)) +
    theme_light(base_size = 6) + xlab('') + ylab('') + ggtitle("Summary matrix") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5 ),
          legend.position = "bottom")


lay <- rbind(c(1,1,2,2),
             c(NA,3,3,NA))
grid.arrange(m1,m2,m3,m4, layout_matrix = matrix(1:4,2,2, byrow = T))

```


### Exponential random graph models
The coefficients in exponential random graph models present the log-odds of the existence of a link given the similarites (Jaccard distances) or matching of node attributes (e.g. scale). Thus, their results are read similarly to those of a logistic regression. Instead of the intercept here we have the `nonzero` term which indicates what the odds are of the existence of a link. Since the networks are weighted by the number of shared drivers, domino effects or hidden feedbacks, the term `sum` accounts for the weight of the link when it exists. We used Jaccard similarity on the attributes coded in the regime shifts database to quantify how similar two regime shifts are and how the similarity affects the odds of a link. Since our hypotheses were related to the temporal and spatial scales at which the regime shifts occurred, we also modeled such terms not only as similarity (a score between 0 and 1), but also as categorical mutually exclusive variables. The original data from the regime shifts database coded for categorical non-exclusive variables. For spatial scale, when two different categories where present (e.g. local, national), we choose the larger category as the unique factor. For temporal scales, regime shifts were often coded as belonging to two categories (e.g. decades and centuries), thus we create a variable `time-range` that bundles both minimum and maximum reported. Spatial and temporal scales are then also modelled as factors too, the term `nodefactor` in each model accounts for how more likely is to have a link when the factor is present, while the `nodematch` term accounts for the homogeneity effect of two nodes sharing the same node attribute (same scale). The analysis was performed in R statistical language [@RCoreTeam:2012wf] using packages for modeling networks with exponential random graph models [@Handcock:2008p5095; @Hunter:2008vh; @Morris:2008ty; @Hunter:2007bq].

\pagebreak

**Table SM1.** Exponential random graph models for \textit{shared drivers}
```{r tableSM2, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}

suppressMessages(stargazer::stargazer(fit.null1, fit.w1, fit.w1a,
    align = F, header = F, type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table SM2. Exponential random graph models for shared drivers (\textit{number of drivers shared})",
    covariate.labels = c("Non-zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provising services", "Regulating services", "Cultural services",
                         "Human wellbeing", "Spatial scale: National",
                         "Spatial scale: sub-continental", "Temporal scale: month-year",
                         "Temporal scale: week-month", "Temporal scale: year-decade", "Spatial scale match",
                         "Temporal scale match", "Spatial scale similarity", "Temporal scale similarity",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Shared drivers",
    add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null1$mle.lik,2),
                       round( fit.w1$mle.lik,2),
                       round(fit.w1a$mle.lik,2))))
    )

```

\pagebreak

**Table SM2.** Exponential random graph models for \textit{domino effects}
```{r tableSM3, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}
suppressMessages(stargazer::stargazer( fit.null2, fit.w2, fit.w2a, fit.w2b,
    align = F, header = F, type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table SM3. Exponential random graph models for \textit{domino effects}",
    covariate.labels = c("Non-zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provising services", "Regulating services", "Cultural services",
                         "Human wellbeing",
                         "Spatial scale similarity", "Temporal scale similarity",
                         "Spatial scale: national", "Spatial scale: sub-continental",
                         "Temporal scale: year", "Temporal scale: month", "Temporal scale: decade",
                         "Indegree when spatial scale national", "Indegree when spatial scale subcontinental",
                         "Indegree when temporal scale year", "Indegree when temporal scale month",
                         "Indegree when temporal scale decade", "Outdegree when spatial scale national",
                         "Outdegree when spatial scale subcontinental", "Outdegree when temporal scale year",
                         "Outdegree when temporal scale month", "Outdegree when temporal scale decade",
                         "Spatial scale match: local", "Spatial scale match: national",
                         "Spatial scale match: sub-continental", "Temporal scale match",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Domino effects",
     add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null2$mle.lik,2),
                       round( fit.w2$mle.lik,2),
                       round(fit.w2a$mle.lik,2),
                       round(fit.w2b$mle.lik,2))))
    )
```

\pagebreak

**Table SM3.** Exponential random graph models for \textit{hidden feedbacks}
```{r tableSM4, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}
suppressMessages(stargazer::stargazer(fit.null3, fit.w3, fit.w3a,
    align = F, header = F,  type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table SM4. Exponential random graph models for \textit{hidden feedbacks}",
    covariate.labels = c("Non-zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provising services", "Regulating services", "Cultural services",
                         "Human wellbeing", "Spatial scale", "Temporal scale",
                         "Spatial scale: national", "Spatial scale: sub-continental",
                         "Temporal scale: year", "Temporal scale: month", "Temporal scale: decade",
                         "Spatial scale match", "Temporal scale match",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Hidden feedbacks",
    add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null3$mle.lik,2),
                       round( fit.w3$mle.lik,2),
                       round(fit.w3a$mle.lik,2))))
    )
```


\pagebreak
