---
title: "Cascading regime shifts within and across scales"
author: |
    | Juan C. Rocha^1,2^, Garry Peterson^1^, Örjan Bodin^1^, Simon A. Levin^1,2,3,4,5^
    | \small ^1^Stockholm Resilience Centre, Stockholm University, Kräftriket 2B, 10691 Stockholm
    | \small ^2^Beijer Institute, Swedish Royal Academy of Sciences, Lilla Frescativägen 4A, 104 05 Stockholm
    | \small ^3^Department of Ecology & Evolutionary Biology, Princeton University, 106A Guyot Hall, Princeton, NJ 08544-1003
    | \small ^4^Arizona State University
    | \small ^5^Resources for the Future
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    latex_engine: pdflatex
    toc: no
  html_document:
    df_print: paged
    toc: no
  word_document:
    toc: no
citecolor: blue
csl: science.csl
documentclass: article
fontsize: 9pt
header-includes:
- \usepackage{dcolumn}
- \usepackage[T1]{fontenc}
- \usepackage{inputenc}
keywords: regime shifts, cascading effects, tele-connections, coupled critical transitions,
  networks
lineno: no
linkcolor: blue
bibliography: cascading.bib
abstract: |
  Regime shifts are large, abrupt and persistent critical transitions in the function and structure of ecosystems. Yet it is unknown how these transitions will interact, whether the occurrence of one will increase the likelihood of another, or simply correlate at distant places. Here we explore two types of cascading effects: domino effects create one-way dependencies, while hidden feedbacks produce two-way interactions. We compare them with the control case of driver sharing which can induce correlations. Using 30 regime shifts described as networks, we show that \textasciitilde{}45% of regime shift pairwise combinations present at least one plausible structural interdependence. The likelihood of cascading effects depends on cross-scale interactions, but differs for each type. Management of regime shifts should account for potential connections.
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf", dpi = 300)
options(knitr.table.format = "latex")
options(tinytex.verbose = TRUE)

set.seed(12345)
library(knitr)
library (dplyr)
library (tidyr)
library(forcats)
library (network)
library (sna)
library(RColorBrewer)
library(ggplot2)
library(GGally)
library(viridis)
library(ggrepel)

# for plotting
library(ggmap)
library(maptools)
library(maps)
library(mapproj)
library(gridBase)
library(grid)
library(gridExtra)
library(png)

## for cleaning models
library(car)
library(broom)
library(sandwich); library(lmtest)

### ERGMS for Regime Shifts
library(ergm.count)

## for tables
library(kableExtra)

## If compiling from R directly:
# setwd('~/Documents/Projects/Cascading Effects/Domino')
# Sys.setenv(RSTUDIO_PANDOC="/Applications/RStudio.app/Contents/MacOS/pandoc")
# rmarkdown::render('170125_draft.Rmd', output_format = 'pdf_document')

  # html_document:
  #   code_folding: hide
  #   dev: pdf
  #   self_contained: yes
  #   toc: yes
  #   toc_float:
  #     collapsed: no
  #     smooth_scroll: no
  # word_document: null

# - \usepackage{lineno}
# - \linenumbers
```


```{r data_ergm, include=FALSE}
# this loads the data, it was cleaned and prepared with script 170329_read_data.R
# and load the ergm models fitted with script 170329_ergms.R
load('~/Documents/Projects/Cascading Effects/data/181001_ergm_data.RData')


## updated functions used for plotting
plotnet <- function(net, ...){
	plot.network(net, #mode='circle',
			vertex.col = alpha(net %v% 'col', 0.7),
			# label = network.vertex.names(net),
			# label.cex= 1,
			main = net %n% 'name', col.main = "gray44", cex.main = 1.2,
			vertex.border = 0,
			usecurve=T,
			vertex.cex= 2, #2 + scale(net %v% 'fb'),
			label.pos= 5,
			edge.col= alpha(net %e% 'col' , 1),
			edge.lwd =  rep(0.01, max(valid.eids(net))), #0.05 + net %e% 'fb',
			edge.curve = 0.01,
			displayisolates=T, pad = 0
			)
}

plotnet2 <- function(net, ...){
	plot.network(net, #mode='circle',
			vertex.col = alpha(net %v% 'col', 0.7),
			# label = network.vertex.names(net),
			# label.cex= 1,
			main = net %n% 'name', col.main = "gray44", cex.main = 1.2,
			vertex.border = 0,
			usecurve=T,
			vertex.cex= 2 + scale(net %v% 'fb'),
			label.pos= 5,
			edge.col= alpha(net %e% 'col' , 1),
			edge.lwd =  0.01 * net %e% 'fb',
			edge.curve = 0.01,
			displayisolates=T, pad = 0 # no need of pad for labels
			)
		}
## A function to create the network, add polarity and cycles / feedbacks as edge attributes
rs.net <- function (dat, i){
	# filter dataset
	dat <- filter(dat, Regime.Shift == levels(dat$Regime.Shift)[i],
	 Polarity == 1 | Polarity == -1)

	# build network
		rs.x <- network(select(dat, Tail, Head, Polarity, col), directed = T, ignore.eval=FALSE, matrix.type = 'edgelist')

	# add cycles to nodes and edges.
		fb.sum <- kcycle.census(
		    rs.x, maxlen=network.size(rs.x),
		    mode='digraph',
		    tabulate.by.vertex=T,
		    cycle.comembership='sum')
		# number of cycles per nodes
		rs.x %v% 'fb' <-  diag(fb.sum$cycle.comemb)
		# number of cycles per edge
		rs.x %e% 'fb' <- as.sociomatrix(rs.x) * fb.sum$cycle.comemb
		#J161106: Note that cycle.comemb is a matrix with comemberships, so two nodes that are not connected in the network can belong to the same cycles. That's why I multiply by the adjacency matrix, to isolate only the edges. cycle.comemb can be used later for discovery of inconvenient feedbacks.
	## add vertex attributes
		rs.x %v% 'col' <- ifelse(colSums(fb.sum$cycle.count)[-1] == 0, "#E41A1C","#8DA0CB") # "#FFD92F"
		rs.x %n% 'name' <- levels(dat$Regime.Shift)[i]
	return(rs.x)
}

```

**Abrupt changes in ecosystems can be connected to others in time and space**

## Introduction

Regime shifts occur across a wide range of social-ecological systems  [@Scheffer:2003p339; @Scheffer:2001uu; @Biggs:2018hx]. They are difficult to predict and reverse [@Boettiger:2013bm; @Hastings:2010p5336], and often produce sustained shifts in the availability of ecosystems services [@Carpenter:2009jr]. When a system undergoes a regime shift it moves from one set of self-reinforcing processes and structures to another [@MAY:1977p3426; @Holling:1973p6861; @Scheffer:2001uu; @Anonymous:2004bq]. Changes in a key variable (e.g. temperature in coral reefs) often make a system more susceptible to shifting regimes when exposed to shock events (e.g. hurricanes) or the action of external drivers (such as fishing) [@Scheffer:2012cta]. Over 30 different regime shifts in social-ecological systems have been documented [@Biggs:2018hx], and similar non-linear dynamics are seen across societies, finance, language, neurological diseases, or climate [@Sole:2011us; @Scheffer:2009wl]. As humans increase their pressure on the planet, regime shifts are likely to occur more often and more severely [@Rocha:2015du; @Drijfhout:2015hj; @Steffen:2018ku].

An emergent challenge for science and practice is that regime shifts can potentially lead to subsequent regime shifts. We define a regime shift as cascading when its occurrence may affect the occurrence of another regime shift. A variety of causal pathways connecting regime shifts have been identified (See Table S1). For example, eutrophication is often reported as a regime shift preceding hypoxia or dead zones in coastal areas [@Diaz:2008p199]. Similarly, hypoxic events have been reported to affect the resilience of coral reefs to warming and other stressors in the tropics [@Altieri:2017bl]. If, why, and how a regime shift somewhere in the world could affect the occurrence of another regime shift remain largely open questions and a key frontier of research [@Liu:2015go; @Hughes:2013cv].

Research on regime shifts is often confined to well defined branches of science, reflecting empirical, theoretical [@Carpenter2003rsi], or predictive approaches [@Scheffer:2012cta; @Scheffer:2009p4449]. These approaches require a deep knowledge of the causal structure of the system or a high quality of spatio-temporal data. Hence research on regime shifts has generally focused on the analysis of individual types of regime shifts rather than potential interactions across systems. This paper takes another approach and instead explores potential cascading effects among a large set of regime shifts. We investigate two types of interconnections: domino effects and hidden feedbacks. Domino effects occur when the feedback processes of one regime shift affects the drivers of another regime shift, creating a one-way dependency [@Rocha:2010vv; @Hughes:2013cv; @Scheffer:2012cta]. A feedback mechanism is a self-amplifying or dampening process characterized by a pathway of causal processes that return to its origin creating a cycle. Two-way interactions rise when two regime shifts combined generate new feedbacks that cannot be identified in the separated regime shifts [@Liu:2015go; @Rocha:2010vv]; and if strong enough, they could amplify or dampen the coupled dynamics. We call them hidden feedbacks because they only show when two regime shift networks are combined. We contrast these cascading effects, in which the occurrence of a regime shift give rise to subsequent regime shifts, with the potentially multiplying albeit different effect of two regime shifts being caused by common drivers. Driver sharing is likely to increase correlation in time or space among regime shifts, but not necessarily interdependence [@Hughes:2013cv; @Rocha:2015du].

## Hypotheses of cascading effects

In analyzing individual regime shifts, bifurcation theory often treats drivers as slow variables, which assumes that their change is relatively slower than changes in variables that describe the state of the system [@Crepin:2006ex; @Kuznetsov:2004vv; @Scheffer:2003p339; @Scheffer:2001uu; @Sole:2011us; @Scheffer:2009wl]. Applying the same logic to pairs of regime shifts, we first expect that domino effects will be dominated by connections from regime shifts occurring at larger spatial scales and slower temporal dynamics than regime shifts receiving the connection [@Dekker:2018jk; @Kinzig:2006p3420]. Secondly, hidden feedbacks are expected to occur when scales match in space and time, because for a new feedback to emerge, regime shifts need to be somewhat aligned in the scale at which their process operates. Thirdly, regime shifts occurring in similar ecosystem types or land uses will be subject to relatively similar sets of drivers, and thus we expect driver sharing to be context specific [@Hughes:2013cv; @Rocha:2015du].

The three hypotheses were tested by analyzing regime shifts as networks of drivers and feedback processes. These directed signed graphs allow us to explore driver co-occurrence, directional pathways, and emergent feedback cycles of coupled regime shift networks [Fig. 1 and @Anonymous:jQuoYC8e for a worked example]. The empirical basis for our investigation draws from the [regime shifts database](www.regimeshifts.org)[@Biggs:2018hx] (Fig S1, S2), which offers syntheses of over $30$ types of regime shifts, and $> 300$ case studies based on literature review of $> 1000$ scientific papers. The database describes regime shifts in terms of their alternative regimes, drivers, feedback mechanisms, impacts on ecosystem services, and management options. It provides a set of $75$ categorical variables about impacts, scales, and evidence types used to test our expectations [Table S2, see @Anonymous:jQuoYC8e for processing of variables]. The database consistently encodes regime shifts into causal diagrams as a graphical summary of the drivers and underlying feedbacks of each regime shift (Fig S1). Causal diagrams for each regime shift were converted into a network by creating the adjacency matrix $A$, where $A_{i,j}$ is $1$ if there is a connection or zero if not [@Anonymous:jQuoYC8e]. A link between two nodes in these networks means that there is at least one scientific paper reviewed in the database providing some evidence for a causal relationship [@Rocha:2015du; @Biggs:2018hx].

Three response variables matrices were created by merging pairs of regime shift networks (Fig 1): (i) for driver sharing it is the number of common drivers, (ii) for domino effects it is the number of directed pathways that connect two regime shifts, and (iii) for hidden feedbacks it is the number of new cycles that emerge when joining two regime shift networks [Fig. 1, @Anonymous:jQuoYC8e]. The hypotheses were tested using exponential random graph models [@Anonymous:jQuoYC8e]. In this framework, our research question can be rephrased as 'What is the likelihood of a link between regime shifts in the response variable matrix, and what features change this likelihood?' As explanatory variables we used the regime shift database categorical variables focusing on how similar two regime shifts are, and whether the similarity increases the likelihood of having a link on the response variable matrices [@Anonymous:jQuoYC8e]. The specification for the model follows a Poisson reference distribution [@Krivitsky:2012uo] given that the response variables contain weighted links of count data --- how many domino effects, hidden feedbacks, or shared drivers link pairs of regime shifts.

```{r Fig1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 3.75, fig.width = 4.75, fig.align = 'center', fig.cap = "Method scheme. Pairs of regime shift causal networks were merged to create a response variable matrix that accounted for drivers shared, domino effects, or hidden feedbacks. In all examples two minimal regime shifts are depicted as causal diagrams, drivers are colored red and variables belonging to feedbacks purple. For driver sharing, the joint network is simplified as a two-mode network which allows us to study the co-occurrence of drivers (in red) across regime shifts (in blue). Driver a is shared by regime shifts 1 and 2 but not driver b. The response variable matrix counts the number of drivers shared by all pairwise combinations of regime shifts. For domino effects, two regime shift networks are joined together where driver c in regime shift 2 is also part of a feedback process in regime shift 1, creating a one-way dependency (orange link) between the two regime shifts. The response variable matrix counts all the one-way causal pathways between pairwise combinations of regime shifts. For hidden feedbacks, two minimal regime shifts when joined together give rise to a new unidentified feedback (orange circular pathway). The response variable matrix counts all hidden feedbacks that arise when merging pairwise combinations of regime shifts. The 30 causal networks used and the labeled matrices of the resulting response variables are shown in Fig. S1 and S3.", dev = "pdf", dpi = 300, dev.args = list(pointsize = 6)}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_Fig1.pdf")

```

## Results

Regime shifts can be structurally interdependent (Fig. 2). The three response variables combined show that ~45% of the regime shift couplings analyzed present structural dependencies in the form of one-way interactions for the domino effect or two-way interactions for hidden feedbacks. While ~5% and ~2% of the couplings present only domino effects and hidden feedbacks, respectively, ~28% of the pairwise combinations are linked through two different types of connections, and ~9% by all three of them. Only for 19% of the pairwise combinations can we be certain with the current dataset that there is no cascading effects. However, the discovery of new drivers or feedback mechanisms underlying these dynamics could reduce this estimate.

```{r Fig2, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 3, fig.width = 4.75, dev.args = list(pointsize= 6), fig.align='center', fig.cap = "Potential structural dependencies between regime shifts. The three response variables combined (A) shows eight different possibilities in which regime shifts can interact through cascading effects. Driver sharing is the most common type found (B). Domino effects and hidden feedbacks alone or in combination account for 45 per cent of all pairwise regime shift couplings analyzed (C) implying structural dependence", dev = "pdf", dpi = 300}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_FigSummary.pdf")

```
Driver sharing is the most common type of connection found (Fig. 2). Regime shifts can correlate in time and space due to common drivers, but they do not necessarily become interdependent[@Hughes:2013cv; @Rocha:2015du]; this is, the occurrence of one does not affect the probability of the second occurring. Of all pairwise regime shift combinations 36% were coupled only by driver sharing. The resulting matrix for driver sharing describes the co-occurrence patterns of `r dim(bipmat)[1]` drivers across the `r dim(bipmat)[2]` regime shifts analyzed. In our sample, aquatic regime shifts tend to have and share more drivers, although the driver sharing is not exclusively with other aquatic regime shifts (Fig. 3A, 4B). The highest driver co-occurrence was found between regime shifts in kelps, marine eutrophication, and the collapse of fisheries. Terrestrial and polar regime shifts tended to have fewer and more specific sets of drivers. Large scale regime shifts in polar and sub-continental areas (e.g. Monsoon weakening) have fewer drivers but are hotspots of sharing, typically including climate related drivers. Drivers that co-occurred most frequently were related to food production, climate change and urbanization, yet none of them is ubiquitous in our sample (Fig. 3B).

In line with our expectations, regime shifts were more likely to share drivers when they occurred in similar land uses, but not necessarily under the same ecosystem types (p << 0.001, Fig. 4, Table S3). We did not expect cross-scale interactions in driver sharing, yet we found that it is more likely to share drivers in dynamics that are faster in time (from weeks to years), and when spatial scales match. Regime shift impacts on ecosystem services and human well-being were related to driver sharing. We found that impacting similar regulating and provisioning services increases the likelihood of common drivers.


```{r Fig3, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 6, fig.width = 4.75, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Patterns of cascading effects. Regime shifts are ranked according to their role in (A) driver sharing, (C) domino effect, and (E) hidden feedbacks. Key variables involved on cascading effects are shown in (B) for driver sharing, (D) domino effects, and (F) hidden feedbacks. The distribution of drivers shared per regime shift (A) with respect to the number of drivers each one has (black points) shows that regime shifts in aquatic environments tend to have and share more drivers. WAIS abbreviates West Antarctica Ice Sheet collapse. Regime shifts that produce most domino effects (high outdegree) are Earth system tipping points, while the regime shifts that receive the most (high indegree) occur in aquatic and land-water interface (C); labels are plotted only for regime shifts where the maximum number of domino effects (4) is found. Most variables associated with domino effects are related to climate and transport mechanisms (D). These variables are part of a feedback mechanism in one regime shift that are in turn drivers in another regime shift. Hidden feedbacks occur typically in terrestrial and Earth system regime shifts. (E) shows the distributions of hidden feedbacks organized by higher to lower mean number of feedbacks. Boxplots are shown in log-scale after zero values have been removed. The variables most often involved in hidden feedbacks have high betweenness and closeness centralities (F) calculated on the network of all regime shifts in our sample (N = 30). These measures reveal the variables (labeled) that lie on most shorter pathways from all other variables in the network.', dev = "pdf", dpi = 300}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_Fig2.pdf")

# Statistical models summary. The response variables are the matrices of regime shifts interconnections for each cascading effect type (Fig 1). The terms fitted correspond to the categorical variables from the regime-shifts database (Table S2) that capture the similarity of regime shifts regarding their impacts on ecosystem services and human well being, the scales at which they occur, reversibility and evidence type (See SM Methods). Results from exponential random graph models are interpreted similarly to those of logistic regression where the non-zero and sum terms are the equivalent of an intercept. The non-zero term is the likelihood of the existence of a link in the response variable network, while the sum corresponds to its weight. Only best models are shown (lowest AIC or BIC), for complete results of null and alternative models fitted see Tables S3, S4 and S5.

```

```{r Fig4, cache = TRUE, echo = FALSE, error=FALSE,  warning=FALSE, message=FALSE, fig.height = 6, fig.width = 4.75, dev.args = list(pointsize= 6), fig.align='center', fig.cap = 'Cascading effects across scales. Summary of the statistical results (A), only models with the lower AIC were included on the figure. Tables S3-S5 complements the figure with alternative models fitted. Circular plots in (B-J) shows the mixing matrices of cascading effects (driver sharing B,C,H, domino effects C,F,I, hidden feedbacks D,G,J) according to ecosystem type (B,C,D), spatial scales (E,F,G), and temporal scales (H,I,J).', dev = "pdf", dpi = 300}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_Fig3.pdf")


```

Evidence of cross-scale interactions for domino effects was only found in time but not in space. As expected, regime shifts producing domino effects have slow temporal dynamics and larger spacial scales. These regime shifts include Earth system tipping elements such as monsoon weakening, thermohaline circulation collapse, and Greenland ice sheet collapse (Fig. 3C). On the other hand, regime shifts influenced by domino effects were often marine and occurred over shorter times and more localized spaces, including mangrove transitions, kelp transitions, and transitions from salt marshes to tidal flats. The statistical models support this observation for temporal scales (Fig. 4, Table S4) but we did not find evidence for spatial ones. Having domino effects was significantly associated with impacting similar regulating and provisioning services (p < 0.001), but the size of the effects are dwarfed by its rarity. The sparse response variable matrix (Fig 1) and the negative coefficient on the sum term in the statistical models (Table S4) show that domino effects are not common. When they do occur it is only through a few pathways between regime shifts (maximum of 4 in our sample). Key variables involved in domino effects were related to climate, nutrients, and water transport (Fig. 3D).

Hidden feedbacks were expected to arise when regime shifts dynamics matched scales in space and time. The statistical analysis supports our hypothesis, regime shifts occurring on the range of decades to centuries and at national scales are more likely to have hidden feedbacks (Fig 4, Table S5). We found fewer hidden feedbacks than one would expect by chance, but when they did occur, they tended to couple regime shifts through multiple feedbacks. Most hidden feedbacks in our sample occurred in terrestrial and Earth systems (Fig 3E, 4). The regime shifts with higher numbers of connections (15 to 18 out of 30 possible) are thermohaline circulation, primary productivity of the Arctic Ocean, forest to savanna, monsoon weakening, and the Greenland ice sheet collapse. Key variables belonging to many hidden feedbacks were related to climate, fires, agriculture and urbanization (Fig 3F).

## Discussion and Conclusions

Regime shifts are ubiquitous in nature, yet how they can interact has reminded an unexplored question. While this question is fundamental for scientists forecasting the dynamics of ecosystems, the answers are relevant for policy makers and managers because regime shifts can impact ecosystem services and human well-being, and hence undermine the achievement of sustainable development goals. Domino effects and hidden feedbacks are often disregarded because research on regime shifts is divided by disciplines that focus on one system at the time. Consequently, data collection and hypothesis testing for coupled systems have largely remained unexplored [@Liu:2015go; @Hughes:2013cv]. While few studies have investigated cascading effects by looking at temperature driven tipping points in the climate systems [@Steffen:2018ku; @Dekker:2018jk; @Gaucherel:2016fm], a growing body of literature has started to offer hypotheses on how different regime shifts can be interconnected (Table S1). Here we have developed a network based method that allow us to systematically identify potential cascading effects, and differentiate whether a regime shift coupling is expected to create structural dependencies in the form of domino effects or hidden feedbacks (Fig. 2).

Our findings align with previous results on the type of variables and processes that can couple regime shifts (See Table S1), highlighting the role of climate, agriculture, and transport mechanisms for nutrients and water (Fig. 3). Recent literature (refs. in Table S1) reports potential linkages between eutrophication and hypoxia, hypoxia and coral transitions, shifts in coral reefs and mangrove transitions, or climate interactions. Other examples in the terrestrial realm report potential increase in Arctic warming from higher fire frequency in boreal forest or permafrost thawing. Regime shifts in the Arctic can impact any temperature driven regime shift in and outside the Arctic [@Peterson:2016ul], including the weakening of the thermohaline circulation. Moisture recycling is a key underlying feedback on the shift from forest to savanna or the Indian monsoon, but also has the potential to couple ecosystems beyond the forest that depend on moisture recycling as an important water source. Changes in moisture recycling can affect mountain forest in the Andes, nutrient cycling in the ocean by affecting sea surface temperature and therefore regime shifts in marine food webs, or exacerbation of dry land related regime shifts. Not all cascading effects reported in the literature and our results are expected to amplify each other. For example, it has been reported that climate tipping points can regulate each other and reduce the probability or regime shifts in forests [@Gaucherel:2016fm; @Brovkin:2009gr].

Guided by the practice of explaining and modeling single regime shifts as emergent dynamics from fast and slow processes [@Crepin:2006ex; @Kuznetsov:2004vv; @Dekker:2018jk; @Carpenter2003rsi; @Holling:1973p6861; @Scheffer:2001uu; @Sole:2011us], we hypothesized that cascading effects between regime shift couplings were determined by cross-scale interactions. For domino effects we only found support of cross-scale interactions in time but not in space. For hidden feedbacks we found evidence of matching in space and time. Together, these results show that the likelihood of regime shift coupling depends on cross-scale interactions, but it differs for each cascading effect type. Lack of evidence for interactions across spatial scales for domino effects, suggests that stochastic and transient dynamics might be playing an  important role in regime shifts [@Hastings:2018gy] and their cascading effects. A major role of stochastic and transient dynamics in regime shifts couplings limits the applicability to early warning signals [@Scheffer:2012cta; @Scheffer:2009p4449] to predict cascading effects [@Dekker:2018jk]. Developing early warning signals for coupled regime shifts is therefore a research need.

Synchronization of regime shifts in time or space is a subject of debate [@Defriez:2017kl; @Frank:2016bo; @Beaugrand:2015by; @Hughes:2013cv]. Temporal correlations ---typically induced by driver sharing--- can be broken by spatial heterogeneity [@Hughes:2013cv], indicating that context matter for correlations to emerge. Spatial heterogeneity can smooth out critical transitions [@Martin:2015kl; @Medeiros:2016vf]. Yet, identifying common drivers is useful for designing management strategies that target bundles of drivers instead of well studied variables independently, increasing the chances that managers will avoid several regime shifts under the influence of the same sets of drivers [@Rocha:2015eea; @Rocha:2015du]. For example, management options for drivers such as sedimentation, nutrient leakage, and fishing can reduce the likelihood of regime shifts such as eutrophication and hypoxia in coastal brackish lagoons, as well as coral transitions in adjacent coral reefs.

Our results complement previous findings (Table S1) by offering a wide spectrum of causal hypotheses about how regime shifts can be coupled. However, the limitations of our method need to be acknowledged. Regime shifts were represented as static networks and the cascading effects were identified by matching two pieces of information: variable names and positions within the causal diagram. Therefore, the method identifies structural dependencies but cannot predict how the dynamics will unfold in space or time. For example, if a connection between mangrove collapse and coral transitions is found through protection against coastal erosion, geographical distance between the two systems or the direction of oceanic currents can change or even cancel out the coupling strength. In fact, coupling strength is expected to change from one place to another. Hence, our method identifies plausible connections between regime shifts, but identifying the conditions that change plausible to probable requires more detailed understanding of regime shift mechanisms. Empirical studies and modeling syntheses are required to translate our identification of possible mechanisms into context sensitive forecasts. Dynamic models of this type of dynamics require careful assumptions about parameter values and the functional form of the system equations. Generalized modeling is a promising technique that does not require particular assumptions, allowing the researcher to reach more general conclusions based on stability properties of the system [@Gross:2009jr; @Lade:2013iwa]. Another potential avenue for future research is looking at how transport mechanisms couple physically distant ecosystems, for example through the moisture recycling feedback [@Keys:2012bo] or international trade [@Liu:2015go]. A key lesson from our study is that regime shifts can be interconnected. Regime shifts should not be studied in isolation under the assumption that they are independent systems. Methods and data collection need to be further developed to account for the possibility of cascading effects.

Our finding that ~45% of regime shift couplings can have structural dependence  suggests that current approaches to environmental management and governance underestimate the likelihood of cascading effects. More attention should be paid to how the Earth is social-ecologically connected [@Liu:2015go], how those connections should be managed, and how to best prepare for regime shifts. Our research suggests that regional ecosystems can be transformed by ecosystem management far away, and conversely, can themselves drive the transformations of other distant ecosystems. Decisions made in one place can undermine the achievement of sustainable development goals in other places. For example, it has been shown that many Arctic regime shifts have the potential to impact non-Arctic ecosystems far away and the provision of their ecosystem services [@Cvijanovic:2017hn; @Peterson:2016ul; @Huang:2017jm]. It implies that whoever does make decisions on management is not necessarily the one that has to deal with the impacts. This issue is evident in governance of water transport systems, whether run-off or atmospheric transport, but it is applicable to other dynamics connecting far away ecosystems through other mechanisms such as climate change, fire, nutrient inputs, or trade. Our results highlight variables that are key for domino effects and hidden feedbacks, e.g. climate, agriculture, transport of nutrients and water. They are also good observables for monitoring early warning indicators of the strengthening of regime shift coupling. How and when nonlinear change can be transmitted across space and time in the Earth system should be considered in assessments and management of future environmental change.


### Acknowledgments
We are grateful to the contributors, reviewers and developers of the regime shifts database. This work was supported by FORMAS grant 942-2015-731 to JR, and National Science Foundation grant OCE-1426746 to SL. JR designed the research, JR and GP curated the data, JR wrote the code and run the analysis with guidance from GP, ÖB and SL; JR, GP, ÖB and SL wrote the paper. Authors declare no competing interests. Data from the regime shifts database is publicly available at [www.regimeshifts.org](www.regimeshifts.org). The version of the database used, and curated causal networks are available in both the regime shifts database and at https://doi.org/10.6084/m9.figshare.7265096.v1. The development version of the code is available at: https://github.com/juanrocha/Domino


\pagebreak


## Supplementary material

This document supplements the article "Cascading regime shifts within and across scales". First we present the materials and methods used in the study. The section is accompanied by a short discussion on the limitations and biases of our dataset, but the interested reader can find a longer discussion presented in [@Biggs:2018hx]. Next, we present a worked example of how the methods work on a pairwise regime shift coupling, how shared drivers, domino effects and hidden feedbacks are identified from their causal diagram representations. Finally, we elaborate on the exponential random graph model results, presenting the motivation for the different model specifications and how they relate to the hypotheses tested in the main article.

Our work is motivated by the question: How regime shifts can be interconnected? Table S1 presents an overview of recent literature that proposes some potential cascading effects.

: Summary of examples from the literature

| Regime shift 1 | Regime Shift 2 | Type | Reference |
|:--------------:|:--------------:|:--------------:|:--------------:|
| Marine eutrophication | Hypoxia | Driver sharing, domino effect, hidden feedback | [@Diaz:2008p199] |
| Hypoxia | Coral transitions | Domino effect | [@Altieri:2017bl] |
| Forest to savanna | Mountain forest transitions | Domino effect | [@Weng:2018hb; @Clark:2015bj; @MoruetaHolme:2015fy] |
| Hypoxia | Marine food webs, fisheries collapse | Domino effect | [@Bakun:2017en; @Weng:2018hb] |
| Forest to savanna | Dry land degradation | Domino effect | [@DebraPCPeters:2004ex] |
| Coniferous to deciduous boreal forest | Arctic sea ice loss | Hidden feedback | [@Young:2016kj; @Kelly:2015iq; @Heon:2014co] |
| Peatland transitions | Arctic sea ice loss | Hidden feedback | [@Zona:2016hq] |
| Arctic sea ice loss | Thermohaline circulation weakening | Hidden feedback | [@Sevellec:2017fk] |
| Antarctic sea ice loss | Marine food webs | Hidden feedback | [@Moore:2018ci] |
| Thermohaline circulation weakening | ENSO | Domino effect | [@Dekker:2018jk] |
| Greenland ice sheet collapse, Arctic sea ice collapse, West Antarctica sea ice collapse | Greenland ice sheet collapse, Arctic sea ice collapse, West Antarctica sea ice collapse | Hidden feedback | [@Steffen:2018ku; @Drijfhout:2015hj; @Gaucherel:2016fm] |
| Greenland ice sheet collapse, Arctic sea ice collapse, West Antarctica sea ice collapse | El Niño Southern Oscillation, Amazon forest to savanna, Thermohaline circulation | Hidden feedbacks | [@Steffen:2018ku; @Gaucherel:2016fm] |
| Greenland ice sheet collapse, Arctic sea ice collapse, West Antarctica sea ice collapse, Thermohaline circulation | East Antarctica ice sheet collapse | Hidden feedbacks | [@Steffen:2018ku; @Drijfhout:2015hj] |
| Fisheries collapse | Marine food webs | Driver sharing, domino effect | [@Tickler:2018gb; @Springer:2018fk; @Moore:2018ci; @Carlson:2018gp; @Branch:2010fn; @Pauly:1998iw] |
| North Atlantic oscillation, Southern oscillation, Pacific decadal oscillation, Global temperature anomaly | North Atlantic oscillation, Southern oscillation, Pacific decadal oscillation, Global temperature anomaly | Hidden feedback, domino effect | [@Seip:2018jm] |
| Arctic sea ice collapse | Dry land degradation | Domino effect | [@Cvijanovic:2017hn] |
| Arctic sea ice collapse | Temperature driven regime shifts | Domino effect | [@Huang:2017jm] |
| El Niño southern oscillation | Forest to savanna, bush encroachment | Domino effect, driver sharing | [@Morton:2017jl; @Hilker:2014hp] |
| Sink to source of methane in wetlands | Greenhouse gas emission driven regime shifts | Domino effect | [@Zhang:2017kr] |
| Contiguous United States summer rainfall | East Asian subtropical monsoon | Domino effect | [@Zhu:2017cn] |
| Greenland sea ice collapse | West African monsoon & dry land degradation | Domino effect | [@Defrance:2017kg] |
| Arctic oscillation | Marine food webs | Driver sharing | [@Beaugrand:2015by]|
| Arctic oscillation | East Asian summer monsoon | Domino effect | [@Chen:2015bs] |
| North Atlantic oscillation | Boreal forest fire regime | Domino effect | [@Drobyshev:2016jr] |
| Arctic sea ice collapse | Bedfast to floating lake ice | Domino effect | [@Alexeev:2016ky] |
| Forest to savanna | Lake food web assemblage | Domino effect | [@Ropke:2017ei]|
| Lake eutrophication | Submerged to floating plants | Domino effect | [@vanGerven:2016jc] |
| Thermohaline circulation | Forest to savanna | Hidden feedback | [@Brovkin:2009gr] |


### Materials and methods:

_Data_: The [regime shifts database](www.regimeshifts.org)  synthesizes scientific literature on our current knowledge of regime shifts. It is an open online repository of regime shift syntheses and case studies. When possible, entries to the regime shift database have been peer reviewed by an expert on the topic to ensure quality and accuracy of its contents [@Biggs:2018hx]. The main data used in this study are the causal loop diagrams described in the database (Fig S1), and a set of 75 categorical variables about regime shift attributes that were used to fit statistical models to explore the role of cross-scale interactions (described below). Out of the 35 regime shifts documented in the database, here we used 30 where complete synthesis exist and causal loop diagrams have been curated.

While Table S1 presents a wide spectrum of potential cascading effects, our study focuses on regime shifts documented in the database. Not all systems with tipping points are included in the database, only systems where the scientific literature has provided evidence of impacts in ecosystem services or human well-being, and evidence of feedback mechanisms that maintain the different regimes or create hysteresis [@Biggs:2018hx]. All examples in Table S1 include systems with tipping points, but not all of them meet the database criteria. Examples of important tipping points that have not been included in the database ---nor our analysis--- include ocean acidification, climate oscillations, or disease outbreaks. They are not described as individual regime shifts in the database yet, but they are found as underlying variables for other regime shifts. For example, ocean acidification is a driver of coral transitions.

The database is based on literature review and as such it inherits the biases of scientific production. For example, variables that are easier to study or to measure are more often reported in papers, or well understood mechanisms are more reported than mechanisms at early stages of confirmation (hypotheses) [@Biggs:2018hx]. Knowledge is heterogenous and the understanding of regime shifts dynamics evolves as new findings are reported. To counterbalance the literature bias, the developers of the regime shift database synthesize regime shift dynamics in causal diagrams (described below), where a link between two variables is not affected by the number of papers reporting the relationship. The database also records whether a driver or feedback mechanism is well established, proposed, or contested, providing a qualitative assessment of the confidence on the existence of the regime shifts and its underlying dynamics [@Biggs:2018hx]. An additional control for the scientific bias is the use of exponential random graph models (described below). This statistical approach provides a framework for dealing with link dependencies; the probability of a link in a network is not independent of the probability of its neighboring links [@Handcock:2008p5095; @Hunter:2008vh; @Morris:2008ty; @Hunter:2007bq]. The bias still exists, but if it affects our results, it would be by neglecting variables seldom reported. The error would be then towards underestimation of potential cascading effects, the main variables involved in cascading effects would remain common across many regime shifts. As noted in the main text, the discovery of new drivers or new feedback mechanisms can reduce our estimate of independent regime shifts couplings (Fig 2).

_Causal loop diagrams (CLDs)_: CLDs represent a collection of causal mechanisms that scientists have reported in their narratives: both empirical (what they choose to sample) or theoretical (what they choose to model). CLDs consist of variables connected by arrows denoting causal influence [@Sterman:2000we], also known as signed directed graphs. Each relationship must have a positive $(+)$ or negative $(-)$ sign that represents the effect of the dependent variable given change in the independent variable [@Lane:2008p6781; @Sterman:2000we]. Although the functional form that underlies each relationship is not necessarily known, positive relationships are proportional while negative ones are inversely proportional _ceteris paribus_. CLDs assume that the causal relationships captured by links are monotonic. Non-linearities are captured by links sets, and thus CLDs do not have self-loops. Feedback loops are the basic structural units of the diagram and emerge by connecting variables in closed directed paths (cycles). Feedback means that once a signal enters the loop, some part of the output is fed back to the input, resulting in amplification or dampening of its own signal. Feedbacks can be reinforcing if the overall polarity of its links is positive, or balancing if negative. Reinforcing feedbacks are usually responsible for behaviors that drive the system out of equilibrium, while balancing feedbacks are responsible for near equilibrium dynamics such as oscillations and delays [@Sterman:2000we]. Note that causal links do not describe the behavior of variables, only the structure of the system: they describe what would happen if there were changes [@Sterman:2000we].

CLDs were curated in the regime shifts database in a way that variables names are consistent, and feedback loops comparable [@Biggs:2018hx]. For example, albedo is the same feedback in the rainforest, Arctic or Antarctic regime shifts. When two variables refer to the same process it is given only one name (e.g. agriculture and cropping = 'agriculture'). When the same variable belongs to qualitatively different causal pathways it is kept as different variables to avoid ambiguities. For example, 'precipitation' refers to the average amount of rain while 'rainfall variability' is the second moment of its distribution; both variables often appear in different causal pathways thus are kept as different variables in our analysis. Fig S1 shows all CLDs used in the analysis [@Biggs:2018hx].


```{r FigS1, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 10, fig.width = 10, dev.args = list(pointsize= 7), fig.align='center', fig.cap = 'Fig S1. Causal networks for all regime shifts analyzed', fig.lp = ""}

## Updated ploting function
# quartz(width = 8, height=6, pointsize= 9)
# detach(package:igraph)
library(network); library(sna)
# quartz(pointsize = 7)
par(mfrow = c(6,5), mar = c(0.25,0.25,1.25,0.25))

for (i in 1:length(levels(dat$Regime.Shift)) ){
	net <- rs.net(dat = dat,  i = i)
	plotnet(net)

}

# quartz.save("Cascading_FigS1_CLDs.png", type = "png", dpi = 800)

```

_Categorical variables:_ We calculated the similarity of each pairwise combination of regime shifts in the database regarding the following categorical attributes (Table S2): *(i)* land use under which the regime shift occurs, *(ii)* ecosystem type; impacts on *(iii)* ecosystem processes, *(iv)* provisioning services, *(v)* regulating services, *(vi)* cultural services, *(vii)* and human wellbeing; as well as *(viii)* the spatial scale at which the regime shift occur, *(ix)* the temporal scales, *(x)* reversibility, and *(xi)* evidence type [@Biggs:2018hx; @Rocha:2015du]. For all $75$ categorical variables encoded, the database reports presence or absence $(0,1)$ allowing us to calculate the Jaccard index and use it as a proxy of how similar two regime shifts are. To facilitate the interpretation of statistical models, the Jaccard distance was rescaled ($x =1 - J_{d}$) so equivalent regime shifts score 1 and completely dissimilar ones score 0. Given that most of our hypotheses relate to the scale at which regime shifts occur, we also modeled scale as a categorical variable to account for matches in the network, not only similarity. Figure S2 shows a map with all generic types of regime shifts and case studies reported in the regime shifts database.



```{r table_2, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}
text_tbl <- data_frame(
    Terms = c("Land use", "Ecosystem type", "Ecosystem processes", "Provisioning services", "Regulating services", "Cultural services", "Impacts on human well-being", "Spatial scale", "Temporal scale", "Reversibility", "Evidence type"),
    Explanation = c("Land use under which the regime shift occurs", "Ecosystem type in which the regime shift occurs", "Impacts on key ecosystem processes",
                    "Impacts on provisioning ecosystem services", "Impacts on regulating ecosystem services", "Impacts on cultural ecosystem services",
                    "Impacts on aspects of human well-being", "Typical spatial scale at which the regime shift occurs",
                    "Typical time scale at which the regime shifts occurs", "Reversibility of the regime shift", "Evidence type in the literature"),
    Variables = c("Urban, small-scale subsistence crop cultivation, large-scale commercial crop cultivation,
                  intensive livestock production (e.g. feedlots, dairies), extensive livestock production (natural rangelands),
                  timber production, fisheries, mining, conservation,
                  tourism, land use impacts are primarily off-site)",
                  "Marine and coastal, freshwater lakes and rivers, temperate and boreal forests, tropical forests, moist savannas and woodlands, drylands and deserts (below ~500mm rainfall/year), mediterranean shrub (e.g. fynbos), grasslands, tundra, rock and ice, agro-ecosystems, planetary",
                  "Soil formation, primary production, nutrient cycling, water cycling, impacts on biodiversity",
                  "Freshwater, food crops, livestock, fisheries, wild animal and plant products, timber, wood fuel, feed fuel and fiber crops, hydropower",
                  "Air quality regulation, climate regulation, water purification, water regulation, regulation of soil erosion, pest and disease regulation, pollination, natural hazard regulation",
                  "Recreation, aesthetic values, knowledge and educational values, spiritual and religious",
                  "Food and nutrition, health (e.g. toxins, disease), livelihoods and economic activity, security of housing and infrastructure, aesthetic and recreational values, cultural identity, social conflict, no direct impact",
                  "Local/landscape (e.g. lake, catchment, community), national (country), sub-continental (e.g. southern Africa, Amazon basin), global",
                  "Weeks, months, years, decades, centuries, unknown",
                  "Irreversible (on 100 year time scale), hysteretic (difficult to reverse), readily reversible, unknown",
                  "Models, paleo-observation, contemporary observations, experiments, other")
)

kable(text_tbl, format = "latex", booktabs = TRUE,
      caption = "Summary of categorical variables") %>%
    kable_styling(full_width = TRUE, latex_options = c("scale_down", "HOLD_position"),
                  font_size = 8) %>%
    column_spec(3, width = "4in") #%>%
    # add_footnote(c("Source: the regime shifts database [www.regimeshifts.org]"), notation = "symbol")
```

```{r FigS2, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 4, fig.width = 6, fig.align = 'center', fig.cap = "Fig S2. Regime shifts around the world. Large points show generic types of regime shifts (n = 35) while small points are case studies (n=324).", dev.args = list(pointsize = 5), fig.lp = ""}

coords <- read.csv2("~/Documents/Projects/Cascading Effects/data/case_coords.csv", dec = '.')

coord_rs <- read.csv(file = '~/Documents/Projects/Cascading Effects/data/rs_coords.csv', dec = '.')
cases <- cbind(cases, coords)
rs <- cbind(rs, coord_rs)
rs$rs <- as.factor(rs$rs)
cases$Type.of.regime.shift <- as.factor(cases$Type.of.regime.shift)
names(cases)[7] <- 'rs'

# correction on RS names for plotting
levels(rs$rs) [17] <- "Marine food webs"
levels(rs$rs) [28] <- "Thermokarst lakes"
levels(cases$rs)[17] <- "Fisheries collapse"
levels(cases$rs)[c(5,22)] <- "Marine food webs"
levels(cases$rs)[c(2,9,12,17,20)] <- "Other"


world <- map_data('world')
world <- fortify(world)

m <- ggplot(world, aes(long, lat)) +
    geom_polygon(fill="grey90", aes(group = group)) +
    geom_path(color="white",aes(group=group), size = 0.2) +
    coord_equal() + theme_void(base_size = 7, base_family = "Helvetica")

m + geom_point(data = cases, aes(lon, lat, color = rs), size = 0.2, show.legend = F) +
    geom_point(data = rs, aes(lon,lat, color = rs), size = 2, show.legend = T ) +
    theme(legend.position = "bottom", legend.text = element_text(size = 4)) +
    scale_color_discrete(guide = guide_legend(title = "Regime shifts documented on the regime shifts database \n Source: www.regimeshifts.org", title.position = 'top', title.hjust = 0.5))

# png(file="rsdb_map.png", width=1280, height = 768, units = "px")
# dev.off()
# quartz.save("rsdb_map.png", width= 12, height=7, pointsize = 10, dpi = 800, bg = "white")

```

_Networks:_ Causal loop diagrams were reinterpreted as networks where node attributes were coded if a node belonged to a feedback loop or not. The latter case is then by definition a driver, an independent variable whose dynamics are not affected by the dynamics of the state variables of the system at hand. Note that networks in ecology usually describe inter-species interactions such as predation or mutualism. Our approach is different from what has been done in ecology. Here a network describes a set of processes, both biotic and abiotic, that can govern ecosystem regime shift dynamics. Our approach is inspired by other network applications to processes such as cell metabolic networks or the network of human diseases, where a process is not captured by a link type (e.g. predation) but by a collection of link interactions (e.g. the Krebs cycle). Thus, individual species are not taken into consideration, but rather their functional roles at the aggregated ecosystem scale (e.g. herbivory).

_Shared drivers:_ To study shared drivers we created a bipartite network of drivers and regime shifts following the method outlined by Rocha et al [@Rocha:2015du]. The statistical models were performed for the case of driver sharing on the matrix of regime shift couples sharing drivers ($A^TA$), and categorical variables from the regime shifts database were used as node attributes or node covariates to test our hypothesis.

_Domino effects:_ Potential domino effects were identified by using the full causal loop diagrams as networks (Fig. S1). The algorithm for identifying domino effects takes the adjacency matrix of two given regime shifts $A_1$ and $A_2$ and identifies all nodes $n \in A_1 \cap A_2$ such that $n$ belong to a feedback in $A_1$ but is a driver in $A_2$. Thus, set differences between causal pathways suggest missing drivers, while set intersection between causal pathways and feedback loop nodes indicate potential domino effects (Fig. 2b). By iterating this simple algorithm we derived how many different pathways exist between every pairwise combination of regime shifts $(N=870)$. The resulting non-symmetrical matrix represents a directed network with regime shifts as nodes and link weights as the number of pathways used for statistical analysis.

_Hidden feedbacks:_ We explored hidden feedbacks by pairwise comparison of causal networks. First, the feedback loops (or k-cycles) are counted by feedback length $k$ for each regime shift matrix  $A_1$ and $A_2$ separately. Then the cycle count is applied to the composite network $A_{1,2}$ of the two regime shifts. The difference between the k-cycles in the composite network $A_{1,2}$ and the k-cycles in the individual networks $A_1$ and $A_2$ constitutes the hidden feedbacks that emerge when the two causal networks are joined. By iterating the same procedure to all pairwise combinations of regime shifts $(N=435)$ we obtained a symmetric directed matrix that was used for the statistical analysis. Although computationally intensive, the search for cycles is feasible in our networks given the small size and relative sparse structures. In fact, the maximum cycle length $k$ is bounded by the size of the network. Out of the `r length(out_inc)` coupled networks analyzed, the maximum feedback length was `r unlist(lapply(out_inc, network.size) ) %>% max`.

```{r matrixes_fig, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 7, dev.args = list(pointsize= 6), fig.align='left', fig.cap = 'Fig S3. Response variable matrices and summary of all cascading effects. (A) shows the controling case of driver sharing, (B) the resulting matrix for domino effects, (C) for hidden feedbacks, and (D) summarizes the results by combining the three matrixes.', dev = "pdf", dpi = 300, fig.lp = ""}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_FigS3.pdf")
```
### A worked example

Figure S4 shows a worked example for each cascading effect analyzed. The upper panel (A) presents causal diagrams as networks for two regime shifts: coral transitions and mangrove transitions. Corals are known to be prone to regime shifts from coral dominated reefs to macro-algae, soft corals, corallimorpharia, sponges, or urchin barrens [@Norstrom:2009p196]. Drivers are variables that are independent of the system dynamics. They affect the state variables e.g. corals, but changes in corals does not change the dynamics of drivers. Thus, nodes with pathways towards state variables but located outside feedbacks are by definition drivers (in red in Fig S4), while variables belonging to circular pathways are denoted as feedback processes (gray in Fig S4). Links are colored blue if the causal relationship is expected to be positive while orange when negative. We did not use link polarity on the analysis. Link polarity is kept in our graphical representations and the archived data because it helps interpreting the diagram, and can be useful for other scientists interested on loop analysis. Corals are a state variable in the coral transitions network, but a driver on the mangrove transition network. Thus it satisfy our definition of domino effect. The orange link on the combined network shows how regime shifts in coral reefs can find a way to impact the likelihood of mangroves also tipping over, through its effects on coastal erosion. In this example, the number of domino effects between coral and mangrove transitions is one recorded on the response variable matrix. Our method identifies the plausible pathway, but it cannot infer the strength of the coupling. Coupling strength will vary from place to place depending on contextual factors such as the direction of currents or distance.

The second example (lower panel) shows a two-way interaction between Arctic summer sea ice loss and coniferous to deciduous forest. The algorithm for finding hidden feedbacks first counts the number of cycles in each network separately. Then it counts the number of cycles on the merged network. Hidden feedbacks are identified when the number of feedbacks in the combined network is higher than the sum of the feedbacks in the separate networks. In our example (B), there are 10 new feedbacks (33 found on the combined network minus 23 found in the two separate networks) that are recorded in the corresponding response variable matrix. Only one hidden feedback is depicted (Fig S4B) of length 12. It describes how changes in fire frequency can change the depth of organic matter, soil temperature and permafrost, ultimately changing the forest composition. Changes in forest composition can have an impact in temperature through albedo, affecting the length of the growing season, and the likelihood of fires.

For the controlling case of driver sharing, we simply count the number of common drivers for all regime shift couplings. In our example (A) there are 9 driver shared: agriculture, climate change, deforestation, greenhouse gases, hurricanes, ocean acidification, sea surface temperature, sedimentation, and urbanization; while in our example (B) there is only one: climate change. These numbers are recorded on the response variable matrix for driver sharing.




```{r FigS4, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 6, fig.width = 7, dev.args = list(pointsize= 7), fig.align='center', fig.cap = 'Fig S4. Identifying cascading effects. A worked example for identification of a domino effect bewteen coral transitions and mangrove transition (A); and a hidden feedback between Arctic sea ice loss and coniferous to deciduous forests (B).', fig.lp = ""}

knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_FigS4.pdf")

```


### Exponential random graph models

This section complements the methods and results from the exponential random graph models. The modeling technique used the number of drivers shared, the number of domino effects and the number of hidden feedback as the three key response variables (Fig 1). The use of exponential random graph models is motivated because they take into account network structure and allow us to control for sample bias. They investigate the odds of the existence and weight of a link and the results are interpreted similarly to a logistic regression.

The coefficients in exponential random graph models present the log-odds of the existence of a link given the similarities (Jaccard distances) or matching of node attributes (e.g. scale). Instead of the intercept here we have the `non zero` term which indicates what the odds are of the existence of a link. Since the networks are weighted by the number of shared drivers, domino effects or hidden feedbacks, the term `sum` accounts for the weight of the link when it exists. We used Jaccard similarity on the attributes coded in the regime shifts database to quantify how similar two regime shifts are and how the similarity affects the odds of a link. Since our hypotheses are related to the temporal and spatial scales at which the regime shifts occurred, we also modeled such terms not only as similarity (a score between 0 and 1), but also as mutually exclusive categorical variables. The original data from the regime shifts database coded for categorical non-exclusive variables. For spatial scale, when two different categories were present (e.g. local, national), the larger category was chosen as unique factor. For temporal scales, regime shifts were often coded as belonging to two categories (e.g. decades and centuries). Thus, a variable `time-range` was created bundling both minimum and maximum reported. Spatial and temporal scales were then also modeled as factors. The term `nodefactor` in each model accounts for how more likely is to have a link when the factor is present, while the `nodematch` term accounts for the homogeneity effect of two nodes sharing the same node attribute (same scale). The analysis was performed in R statistical language [@RCoreTeam:2012wf] using packages for modeling networks with exponential random graph models [@Handcock:2008p5095; @Hunter:2008vh; @Morris:2008ty; @Hunter:2007bq].

Different statistical models were fitted to test our hypotheses. In all cases, a null model with only `non zero` and `sum` terms was fitted. It is a reference point that allows model selection. If alternative models do not improve the Akaike information criterion, or the Bayesian information criterion, the null model is the best at explaining the patterns in the data. The interpretation would be then that the processes underlying the formation of our response variable matrices are random. We fitted two additional models for the case of driver sharing (Table S3), one looking at spatial and temporal scales as factors (model 2), and another using the Jaccard similarity (model 3). Model 2 was the best model (lower AIC) reported in Fig 4.

For the case of domino effects we compared the null model with three additional model specifications (Table S4). Model 2 tests for cross-scale interactions treating scale as a similarity covariant, while model 3 tests for scale as individual factors with an additional term for matching scales. Model 4 takes into account the directionality of the domino effects by treating scale as a factor term in the case of outdegree which produces the domino effect, as opposed to indegree which receives the domino effect. This specification was necessary because the response matrix for domino effects is asymmetric, directionality matters. Note that assessing model performance with AIC suggests model 2 is the best model reported in Fig 4.

For the case of hidden feedbacks we compared the null model with two additional models (Table S5). Model 2 treats scales as a similarity covariant while model 3 treats them as categorical factors and includes the matching categories term. Model 3 is then the model that explicitly tests our hypothesis of scale match, and also the model that better explains the patterns on our data (lower AIC) reported in Fig 4.

### Notes on reversibility and evidence type

All models showed that reversibility and evidence type were significant predictors of our response variables. Higher similarity in reversibility increases the likelihood of domino effects and driver sharing, while lower similarity in reversibility increases the likelihood of hidden feedbacks. On the other hand, higher similarity on evidence type increases the likelihood of hidden feedbacks, while low similarity of evidence type increases the likelihood of domino effects. Reversibility is an important aspect of regime shifts. However, we did not elaborate further because the diversity of reversibility options is rather low. In our sample of 30 regime shifts, 9 are reported with hysteresis, for 6 reversibility is unknown, and for 5 it is reported both irreversible (on a 100 year time scale) or hysteresis (Fig S5). Reversibility and evidence type a non-exclusive categorical variables where most of the regime shifts in our sample have the same categories, so they have high similarity in any case. Contrary to other categorical variables in our models, reported evidence type or reversibility might change as knowledge about regime shifts improve. For this reason we did not emphasized this result. We believe it corresponds more to how well studied certain regime shifts are (the scientific bias discussed in the description of our data) rather than fundamental patterns of how regime shifts can interact. For example, the fact that the coefficients flip in the case of driver sharing compared to hidden feedbacks could be explained by driver sharing occurring more often between (better documented) aquatic regime shifts, while hidden feedbacks are more predominant between terrestrial and Earth system regime shifts (which are less well understood).

```{r FigS5, cache = TRUE, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.height = 2, fig.width = 7, dev.args = list(pointsize= 7), fig.align='center', fig.cap = 'Fig S5. Reversibility and evidence type.', fig.lp = ""}

 knitr::include_graphics(path = "~/Documents/Projects/Cascading Effects/Figures/Cascading_FigS5.pdf")

```

\pagebreak

**Table S3.** Exponential random graph models for \textit{shared drivers}
```{r tableSM2, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}

suppressMessages(stargazer::stargazer(fit.null1, fit.w1, fit.w1a,
    align = F, header = F, type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table S2. Exponential random graph models for shared drivers (\textit{number of drivers shared})",
    covariate.labels = c("Non zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provisioning services", "Regulating services", "Cultural services",
                         "Human well-being", "Spatial scale: national",
                         "Spatial scale: sub-continental", "Temporal scale: month-year",
                         "Temporal scale: week-month", "Temporal scale: year-decade", "Spatial scale match",
                         "Temporal scale match", "Spatial scale similarity", "Temporal scale similarity",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Shared drivers",
    add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null1$mle.lik,2),
                       round( fit.w1$mle.lik,2),
                       round(fit.w1a$mle.lik,2))))
    )

```

\pagebreak

**Table S4.** Exponential random graph models for \textit{domino effects}
```{r tableSM3, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}

suppressMessages(stargazer::stargazer( fit.null2, fit.w2, fit.w2a, fit.w2b,
    align = F, header = F, type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table S3. Exponential random graph models for \textit{domino effects}",
    covariate.labels = c("Non zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provisioning services", "Regulating services", "Cultural services",
                         "Human well-being",
                         "Spatial scale similarity", "Temporal scale similarity",
                         "Spatial scale: national", "Spatial scale: sub-continental",
                         "Temporal scale: year", "Temporal scale: month", "Temporal scale: decade", "Spatial scale match",
                         "Indegree when spatial scale national", "Indegree when spatial scale subcontinental",
                         "Indegree when temporal scale year", "Indegree when temporal scale month",
                         "Indegree when temporal scale decade", "Outdegree when spatial scale national",
                         "Outdegree when spatial scale subcontinental", "Outdegree when temporal scale year",
                         "Outdegree when temporal scale month", "Outdegree when temporal scale decade",
                         "Spatial scale match: local", "Spatial scale match: national",
                         "Spatial scale match: sub-continental", "Temporal scale match",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Domino effects",
     add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null2$mle.lik,2),
                       round( fit.w2$mle.lik,2),
                       round(fit.w2a$mle.lik,2),
                       round(fit.w2b$mle.lik,2))))
    )
```

\pagebreak

**Table S5.** Exponential random graph models for \textit{hidden feedbacks}
```{r tableSM4, message = FALSE, warning = FALSE, echo = FALSE, results='asis'}

suppressMessages(stargazer::stargazer(fit.null3, fit.w3, fit.w3a,
    align = F, header = F,  type = 'latex', digits = 2,
    float = F, no.space = T, single.row = T, font.size = "footnotesize",
    title = "Table S4. Exponential random graph models for \textit{hidden feedbacks}",
    covariate.labels = c("Non zero", "Sum", "Land use", "Ecosystem type", "Ecosystem processes",
                         "Provising services", "Regulating services", "Cultural services",
                         "Human wellbeing", "Spatial scale", "Temporal scale",
                         "Spatial scale: national", "Spatial scale: sub-continental",
                         "Temporal scale: year", "Temporal scale: month", "Temporal scale: decade",
                         "Spatial scale match", "Temporal scale match",
                         "Reversibility",
                         "Evidence type"),
    dep.var.labels = "Hidden feedbacks",
    add.lines = list(c("Maximum likelihood estimation",
                       round(fit.null3$mle.lik,2),
                       round( fit.w3$mle.lik,2),
                       round(fit.w3a$mle.lik,2))))
    )
```


\pagebreak
